### 2020-06-25 tex/paper.tex
Start working on the paper
Some headings is a good start!

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T2A]{fontenc}

\title{Compiling refinement types to dependent types}

\begin{document}

\maketitle

\section{Introduction}

\section{Overview}

\section{Calculus definition}

\subsection{Surface language}

\subsection{Core language}

\section{Translation}

\end{document}

### 2020-06-25 tex/build.sh
Start playing around with ott/ottalt

@@ -1,0 +1,4 @@
+#! /bin/sh
+
+ott -tex_wrap false -i surface.ott -o surface.ott.tex
+pdflatex paper.tex

### 2020-06-26 toy/src/Toy/Language/Solver/QueryGen.hs
This is the right way to go wrt variables representation

@@ -36,8 +36,6 @@ type MonadQ m = MonadState RefAnnState m
 propagateRefinements :: MonadQ m => TypedTerm -> m RefAnnTerm
 propagateRefinements (TName ty name) = do
   v' <- freshRefVar
-  -- We could've avoided generating a fresh var and having to insert the extra conjunct,
-  -- but it's perhaps safer to do so since it simplifies a later check that all refinement variables are unique.
   let specedRef = specRefinement v' $ tyRefinement ty
   let refinement | isBaseTy ty = addConjunct (AR $ tv v' |=| tv name) specedRef
                  | otherwise = specedRef

### 2020-06-26 toy/src/Toy/Language/Parser/Util.hs
Support line comments

@@ -49,6 +49,7 @@ parens :: ToyMonad e s m => m a -> m a
 parens = between (lsymbol "(") (lsymbol ")")
 
 lexSpace :: ToyMonad e s m => m ()
-lexSpace = ML.space myspace empty empty
+lexSpace = ML.space myspace lineComments empty
   where
     myspace = void $ takeWhile1P (Just "white space") $ \c -> c /= '\n' && isSpace c
+    lineComments = skipLineComment "--"

### 2020-06-26 toy/src/Toy/Language/Parser/Util.hs
Also skip block comments

@@ -49,7 +49,8 @@ parens :: ToyMonad e s m => m a -> m a
 parens = between (lsymbol "(") (lsymbol ")")
 
 lexSpace :: ToyMonad e s m => m ()
-lexSpace = ML.space myspace lineComments empty
+lexSpace = ML.space myspace lineComments blockComments
   where
     myspace = void $ takeWhile1P (Just "white space") $ \c -> c /= '\n' && isSpace c
     lineComments = skipLineComment "--"
+    blockComments = skipBlockComment "{-" "-}"

### 2020-06-26 toy/test/T00TypesParserSpec.hs
Add a few comments tests

@@ -80,6 +80,11 @@ spec = do
     it "parses pi-bound nested refinement types in pi-bound parens" $
       p "(f : (x : { v : Int | v <= len arr }) -> { v : Int | v > 0 }) -> { v : Int | v >= 0 & v < x }"
         ~~> "f".:("x".:intLeqLenArr ->> intGe0) ->> intBetween0andX
+  describe "Parsing comments" $ let p = parse' ty in do
+    it "ignores line comments" $
+      p "Bool -> Int -- this is a type yay" ~~> bool --> int
+    it "ignores block comments" $
+      p "Bool {- this is something -} -> Int {- smth else -}" ~~> bool --> int
 
 -- Some helpers to make tests a tad more pleasant
 infixr 0 -->

### 2020-06-26 toy/src/Idris/IdeModeClient.hs
Only remove the idris files if they do exist

@@ -27,6 +27,7 @@ import Control.Concurrent.STM
 import Control.Exception(bracket)
 import Control.Monad
 import Control.Monad.Catch(MonadMask, finally)
+import Control.Monad.Extra
 import Control.Monad.IO.Class
 import Control.Monad.Operational
 import Data.Default
@@ -156,8 +157,10 @@ runIdrisClientInst ih@(IdrisInstance (idrStdin, idrStdout, _, _)) = viewT >=> go
 
 cleanupIdrisFiles :: MonadIO m => FilePath -> m ()
 cleanupIdrisFiles path = liftIO $ do
-  removeFile $ "build/ttc/" <> path -<.> "ttc"
-  removeFile $ "build/ttc/" <> path -<.> "ttm"
+  removeFile' $ "build/ttc/" <> path -<.> "ttc"
+  removeFile' $ "build/ttc/" <> path -<.> "ttm"
+  where
+    removeFile' p = whenM (doesFileExist p) $ removeFile p
 
 fmtLength :: Command -> String
 fmtLength cmd = replicate (6 - length hex) '0' <> hex

### 2020-06-26 toy/src/Toy/Language/EnvironmentUtils.hs
argTypes isn't used anywhere but produces a ton of shadowing warnings

@@ -5,7 +5,6 @@ module Toy.Language.EnvironmentUtils
 , buildCombinedMapping
 , buildTypesMapping
 , retType
-, argTypes
 
 , ArgTypes
 , Var2Ty
@@ -27,9 +26,6 @@ funTypesMapping sig def = (arg2type, resType)
   where
     (argTypes, resType) = splitTypes sig
     arg2type = zip (funArgs def) argTypes
-
-argTypes :: FunSig -> [Ty]
-argTypes = fst . splitTypes
 
 retType :: FunSig -> RefinedBaseTy
 retType = snd . splitTypes

### 2020-06-27 toy/app/Main.hs
Start working on porting the compiler

@@ -24,5 +24,5 @@ main = do
           contents <- readFile filename
           case parseFunDecls contents of
                Left err -> putStrLn $ "Parse error:\n" <> err
-               Right ds -> putStr $ compileDecls ds
+               Right ds -> compileDecls ds >>= putStr
        _ -> putStrLn $ "Unrecognized arguments, try running `" <> progName <> " help`"

### 2020-06-27 toy/src/Toy/Language/Compiler.hs

@@ -7,6 +7,7 @@ module Toy.Language.Compiler
 ) where
 
 import qualified Data.HashMap.Strict as HM
+import Control.Monad
 import Data.List
 import Data.Maybe
 import Data.String.Interpolate
@@ -14,16 +15,26 @@ import Data.String.Interpolate
 import Misc.Util
 import Toy.Language.BasicTC
 import Toy.Language.EnvironmentUtils
+import Toy.Language.Solver.QueryGen
+import Toy.Language.Solver.QueryInterp
+import Toy.Language.Solver.Types
 import Toy.Language.Syntax
 
-compileDecl :: [FunSig] -> Decl -> String
-compileDecl ctx Decl { .. } = sig' <> fromMaybe "" def'
+compileDecl :: [FunSig] -> Decl -> IO String
+compileDecl ctx Decl { .. }
+  | Just def <- declDef = do
+      let typedDef = annotateFunDef ctx declSig def
+      let qdef = genQueriesFunDef declSig typedDef
+      (res, solvedFunDef) <- solveDef declSig qdef
+      case res of
+           Correct -> pure $ sig' <> compileFunDef declSig solvedFunDef
+           _ -> error "solve failed"
+  | otherwise = pure sig'
   where
     sig' = compileFunSig declSig <> "\n"
-    def' = (<> "\n") . compileFunDef declSig . annotateFunDef ctx declSig <$> declDef
 
-compileDecls :: [Decl] -> String
-compileDecls decls = unlines $ zipWith compileDecl (inits sigs) decls
+compileDecls :: [Decl] -> IO String
+compileDecls decls = unlines <$> zipWithM compileDecl (inits sigs) decls
   where
     sigs = declSig <$> decls
 
@@ -77,12 +88,12 @@ compileBaseTy TInt = "Int"
 compileBaseTy TIntList = "List Int"
 
 -- TODO need the surrounding context
-compileFunDef :: FunSig -> TypedFunDef -> String
+compileFunDef :: FunSig -> VCFunDef SolveRes -> String
 compileFunDef sig def@FunDef { .. } = [i|#{funName} #{unwords funArgsNames} = #{funBodyStr}|]
   where
     funArgsNames = getName <$> funArgs
     (argTypes, resType) = funTypesMapping sig def
-    funBodyStr = wrapping ctx (TyBase resType) $ unwrapping ctx funBody
[…]

### 2020-06-27 toy/test/T30TermsCompilerSpec.hs

@@ -8,21 +8,24 @@ import Test.Hspec
 import Idris.IdeModeClient
 import Toy.Language.BasicTC
 import Toy.Language.Compiler
-import Toy.Language.Solver
+import Toy.Language.Solver.QueryGen
+import Toy.Language.Solver.QueryInterp
+import Toy.Language.Solver.Types
 
 import TestUtils
 
 checkIdris :: String -> IdrisHandle -> Expectation
 checkIdris program ih = do
-  (ctx, (funSig, funDef)) <- testParseFunWithCtx program
-  let typedFunDef = annotateFunDef ctx funSig funDef
-  res <- solve (buildCtx ctx) funSig typedFunDef
+  (ctx, (sig, def)) <- testParseFunWithCtx program
+  let typedDef = annotateFunDef ctx sig def
+  let qdef = genQueriesFunDef sig typedDef
+  (res, solvedFunDef) <- solveDef sig qdef
   res `shouldBe` Correct
   runIdrisClient ih $ withFile $ \file -> do
     writePrelude file
     mapM_ (write file . compileFunSig) ctx
-    write file $ compileFunSig funSig
-    write file $ compileFunDef funSig typedFunDef
+    write file $ compileFunSig sig
+    write file $ compileFunDef sig solvedFunDef
     testIdrisFile file
 
 spec :: Spec

### 2020-06-27 toy/src/Toy/Language/Compiler.hs
Get rid of the old solver

@@ -15,9 +15,7 @@ import Data.String.Interpolate
 import Misc.Util
 import Toy.Language.BasicTC
 import Toy.Language.EnvironmentUtils
-import Toy.Language.Solver.QueryGen
-import Toy.Language.Solver.QueryInterp
-import Toy.Language.Solver.Types
+import Toy.Language.Solver
 import Toy.Language.Syntax
 
 compileDecl :: [FunSig] -> Decl -> IO String

### 2020-06-27 toy/src/Toy/Language/Solver.hs

@@ -1,226 +1,5 @@
-{-# LANGUAGE GeneralizedNewtypeDeriving #-}
-{-# LANGUAGE FlexibleContexts #-}
-{-# LANGUAGE RecordWildCards, QuasiQuotes, BlockArguments, TupleSections #-}
+module Toy.Language.Solver (module X) where
 
-module Toy.Language.Solver
-( solve
-, SolveContext
-, SolveRes(..)
-
-, buildCtx
-) where
-
-import qualified Data.HashMap.Strict as HM
-import Data.Maybe
-import Data.String.Interpolate
-import Control.Monad
-import Control.Monad.Reader
-import Z3.Monad
-
-import Toy.Language.EnvironmentUtils
-import Toy.Language.Solver.Types
-import Toy.Language.Syntax
-
-newtype SolveContext = SolveContext
-  { visibleSigs :: [FunSig]
-  } deriving (Eq, Ord, Show, Semigroup, Monoid)
-
-buildCtx :: [FunSig] -> SolveContext
-buildCtx = SolveContext
-
-solve :: SolveContext -> FunSig -> TypedFunDef -> IO SolveRes
-solve ctx sig def = undefined {- evalZ3 $ mkScript ctx arg2type resType (funBody def)
-  where
-    (arg2type, resType) = funTypesMapping sig def
-
-newtype Z3VarName = Z3VarName { getZ3VarName :: AST }
-
-newtype SolveEnvironment = SolveEnvironment
-  { z3args :: HM.HashMap VarName (Ty, Z3VarName)
-  }
-
--- This expects that the pi-binders names in the type are equal to the argument names in the definition.
--- TODO explicitly check for this.
-mkScript :: SolveContext -> ArgTypes -> RefinedBaseTy -> TypedTerm -> Z3 SolveRes
-mkScript ctx args target term = do
-  solveEnv <- buildSolveEnv (visibleSigs ctx) args
-
-  convertZ3Result <$> flip runReaderT solveEnv do
-    argsPresup <- genArgsPresup
-
-    res <- Z3VarName <$> mkFreshTypedVar (baseType target) "_res$"
-    resConcl <- genRefinementCstrs target res >>= mkAnd
-
-    TermsCstrs { .. } <- genTermsCstrs res term
-    case mandatoryCstrs of
-         Just cstrs -> assert cstrs
-         Nothing -> pure ()
[…]

### 2020-06-27 toy/test/T20TCSpec.hs

@@ -6,9 +6,7 @@ import Data.String.Interpolate
 import Test.Hspec
 
 import Toy.Language.BasicTC
-import Toy.Language.Solver.QueryGen
-import Toy.Language.Solver.QueryInterp
-import Toy.Language.Solver.Types
+import Toy.Language.Solver
 
 import TestUtils
 

### 2020-06-28 toy/src/Toy/Language/Syntax/Types.hs
Refinements should allow arbitrary Term parametrizations

@@ -16,13 +16,17 @@ import Toy.Language.Syntax.Terms
 data BaseTy = TBool | TInt | TIntList
   deriving (Eq, Ord, Show, Enum, Bounded, Data)
 
-newtype AtomicRefinement = AR { getARTerm :: Term }
+newtype AtomicRefinementT a = AR { getARTerm :: TermT a }
   deriving (Eq, Ord, Show, Data)
 
-data Refinement = Refinement
+type AtomicRefinement = AtomicRefinementT ()
+
+data RefinementT a = Refinement
   { subjectVar :: VarName
-  , conjuncts :: [AtomicRefinement]
+  , conjuncts :: [AtomicRefinementT a]
   } deriving (Eq, Ord, Show, Data)
+
+type Refinement = RefinementT ()
 
 trueRefinement :: Maybe Refinement
 trueRefinement = Nothing

### 2020-06-28 toy/src/Toy/Language/Solver/QueryInterp.hs
Get rid of the orphan instance

@@ -118,10 +118,6 @@ createVar ty varName = do
 getVar :: MonadConvert m => VarName -> m Z3Var
 getVar varName = gets $ (HM.! varName) . variables
 
-instance MonadZ3 m => MonadZ3 (StateT s m) where
-  getSolver = StateT $ \st -> fmap (\solv -> (solv, st)) getSolver
-  getContext = StateT $ \st -> fmap (\ctx -> (ctx, st)) getContext
-
 mkAnd' :: (Foldable t, MonadZ3 m) => t AST -> m AST
 mkAnd' = mk . toList
   where

### 2020-06-28 toy/src/Toy/Language/Compiler.hs
Reimplement `compileAR` in terms of `compileTerm`

@@ -63,22 +63,9 @@ compileRefinement ctx refinement =
        conjs -> "(" <> intercalate ", " (compileAR ctx <$> conjs) <> ")"
 
 compileAR :: Var2Ty -> AtomicRefinement -> String
-compileAR ctx (AR term) = undefined {- [i|v #{opStr} #{argStr} = True|] TODO
+compileAR ctx (AR term) = [i|#{compileTerm ctx typedTerm} = True|]
   where
-    opStr = case op of
-                 ROpLt -> "<"
-                 ROpLeq -> "<="
-                 ROpEq -> "=="
-                 ROpNEq -> "/="
-                 ROpGt -> ">"
-                 ROpGeq -> ">="
-    argStr = case arg of
-                  RArgInt n -> show n
-                  RArgVar var -> unwrap var
-                  RArgVarLen var -> "intLength " <> unwrap var
-    unwrap var | Just varTy <- HM.lookup var ctx
-               , isJust $ tyRefinement varTy = [i|fst #{getName var}|]
-               | otherwise = getName var -}
+    typedTerm = runReader (annotateTypes term) ctx
 
 compileBaseTy :: BaseTy -> String
 compileBaseTy TBool = "Bool"

### 2020-06-28 toy/src/Toy/Language/Compiler.hs
Build fix

@@ -8,6 +8,7 @@ module Toy.Language.Compiler
 
 import qualified Data.HashMap.Strict as HM
 import Control.Monad
+import Control.Monad.Reader
 import Data.List
 import Data.Maybe
 import Data.String.Interpolate

### 2020-06-29 toy/src/Toy/Language/Compiler.hs
Pass the typing context to the refinements

@@ -45,7 +45,7 @@ compileTy :: Ty -> String
 compileTy = go mempty
   where
     go ctx (TyBase RefinedBaseTy { .. })
-      | Just refinement <- baseTyRefinement = [i|(v : #{compileBaseTy baseType} ** #{compileRefinement ctx refinement})|]
+      | Just refinement <- baseTyRefinement = [i|(v : #{compileBaseTy baseType} ** #{compileRefinement ctx baseType refinement})|]
       | otherwise = compileBaseTy baseType
     go ctx (TyArrow ArrowTy { .. })
       | isBaseTy domTy || isJust piVarName = [i|#{lhs} -> #{go ctx' codTy}|]
@@ -56,12 +56,14 @@ compileTy = go mempty
         lhs | Just name <- piVarName = [i|(#{getName name} : #{go ctx domTy})|]
             | otherwise = compileTy domTy
 
-compileRefinement :: Var2Ty -> Refinement -> String
-compileRefinement ctx refinement =
+compileRefinement :: Var2Ty -> BaseTy -> Refinement -> String
+compileRefinement ctx baseTy refinement =
   case conjuncts refinement of
        [] -> "()"
-       [conj] -> compileAR ctx conj
-       conjs -> "(" <> intercalate ", " (compileAR ctx <$> conjs) <> ")"
+       [conj] -> compileAR ctx' conj
+       conjs -> "(" <> intercalate ", " (compileAR ctx' <$> conjs) <> ")"
+  where
+    ctx' = HM.insert (subjectVar refinement) (TyBase $ RefinedBaseTy baseTy Nothing) ctx
 
 compileAR :: Var2Ty -> AtomicRefinement -> String
 compileAR ctx (AR term) = [i|#{compileTerm ctx typedTerm} = True|]
@@ -100,7 +102,8 @@ unwrapping ctx t = unwrapStr (annotation t) $ compileTerm ctx t
 
 wrapping :: Var2Ty -> Ty -> String -> String
 wrapping ctx ty str
-  | Just refinement <- tyRefinement ty = [i|MkDPair {p = \\v => #{compileRefinement ctx refinement}} #{parens str} smt|]
+  | TyBase RefinedBaseTy { .. } <- ty
+  , Just refinement <- baseTyRefinement = [i|MkDPair {p = \\v => #{compileRefinement ctx baseType refinement}} #{parens str} smt|]
   | otherwise = str
 
 unwrapStr :: Ty -> String -> String

### 2020-06-29 toy/test/T05TypesCompilerSpec.hs
Add a couple of simpler tests for the types compiler

@@ -45,6 +45,8 @@ spec = testWithIdris $ do
       it "refinements" $ checkIdris "someFun : { v : Int | v > 0 } -> Bool"
       it "pi-bound vars" $ checkIdris "someFun : (x : Int) -> Bool"
       it "pi-bound vars and refinements" $ checkIdris "someFun : (x : { v : Int | v > 0 }) -> Bool"
+      it "pi-bound vars referred in result" $ checkIdris "someFun : (x : Int) -> { v : Int | v >= 0 & v < x }"
+      it "pi-bound vars referred in another refinement " $ checkIdris "someFun : (x : Int) -> { v : Int | v >= 0 & v < x } -> Int"
       it "pi-bound vars some more" $ checkIdris "someFun : (ls : IntList) -> { v : Int | v >= 0 & v < len ls } -> Int"
       it "pi-bound unrefined vars mentioned in subsequent refinements" $
         checkIdris "add : (x : Int) -> (y : Int) -> { v : Int | v >= x & v >= y }"

### 2020-06-29 toy/src/Toy/Language/Compiler.hs
Don't lose the context when compiling domTy

@@ -54,7 +54,7 @@ compileTy = go mempty
         ctx' | Just varName <- piVarName = HM.insert varName domTy ctx
              | otherwise = ctx
         lhs | Just name <- piVarName = [i|(#{getName name} : #{go ctx domTy})|]
-            | otherwise = compileTy domTy
+            | otherwise = go ctx domTy
 
 compileRefinement :: Var2Ty -> BaseTy -> Refinement -> String
 compileRefinement ctx baseTy refinement =

### 2020-06-29 toy/src/Toy/Language/Syntax/Types.hs
Some more pretty-printer fixes
I shall probably change my IDE font.

@@ -86,8 +86,8 @@ instance Pretty Ty where
     | Just Refinement { .. } <- baseTyRefinement = [i|{ #{getName subjectVar} : #{baseType} | #{pretty conjuncts} } |]
     | otherwise = show baseType
   pretty (TyArrow ArrowTy { .. })
-    | Just piVar <- piVarName = [i|(#{getName piVar} : #{parenPretty domTy}) -> #[pretty codTy}|]
-    | otherwise = [i|#{parenPretty domTy} -> #[pretty codTy}|]
+    | Just piVar <- piVarName = [i|(#{getName piVar} : #{parenPretty domTy}) -> #{pretty codTy}|]
+    | otherwise = [i|#{parenPretty domTy} -> #{pretty codTy}|]
     where
       parenPretty ty | isArrowTy ty = "(" <> pretty ty <> ")"
                      | otherwise = pretty ty

### 2020-06-29 toy/test/T05TypesCompilerSpec.hs
Simplify the test mentioning `len`

@@ -47,7 +47,8 @@ spec = testWithIdris $ do
       it "pi-bound vars and refinements" $ checkIdris "someFun : (x : { v : Int | v > 0 }) -> Bool"
       it "pi-bound vars referred in result" $ checkIdris "someFun : (x : Int) -> { v : Int | v >= 0 & v < x }"
       it "pi-bound vars referred in another refinement " $ checkIdris "someFun : (x : Int) -> { v : Int | v >= 0 & v < x } -> Int"
-      it "pi-bound vars some more" $ checkIdris "someFun : (ls : IntList) -> { v : Int | v >= 0 & v < len ls } -> Int"
+      it "pi-bound vars some more" $
+        checkIdris "someFun : (len : IntList -> Int) -> (ls : IntList) -> { v : Int | v >= 0 & v < len ls } -> Int"
       it "pi-bound unrefined vars mentioned in subsequent refinements" $
         checkIdris "add : (x : Int) -> (y : Int) -> { v : Int | v >= x & v >= y }"
       it "pi-bound vars with refinements referring refinements" $

### 2020-06-29 toy/test/T20TCSpec.hs
Fix one more test with fun substitution so it now passes

@@ -225,12 +225,12 @@ spec = do
            fun : { v : Int | v > 0 } -> { v : Int | v > 0 } -> { v : Int | v >= 0 }
            fun x y = g (f x) (h y)
           |]
-    xit "accepts yet more complicated function application" $ expectSolverOn Correct
+    it "accepts yet more complicated function application" $ expectSolverOn Correct
         [i|
            f : (x : { v : Int | v >= 0 }) -> { v : Int | v > x }
            g : (x : Int) -> { v : Int | v > 0 }
 
-           fun : (x : { v : Int | v > 0 }) -> { v : Int | v >= x }
+           fun : (x : { v : Int | v > 0 }) -> { v : Int | v >= 0 }
            fun x = g (f x)
           |]
     xit "accepts yet more complicated function application" $ expectSolverOn Correct

### 2020-06-29 toy/test/T20TCSpec.hs
Add a slightly more complicated test for functional substitution

@@ -236,6 +236,14 @@ spec = do
     xit "accepts yet more complicated function application" $ expectSolverOn Correct
         [i|
            f : (x : { v : Int | v >= 0 }) -> { v : Int | v > x }
+           g : (x : Int) -> { v : Int | v > x }
+
+           fun : (x : { v : Int | v > 0 }) -> { v : Int | v >= x }
+           fun x = g (f x)
+          |]
+    xit "accepts yet more complicated function application" $ expectSolverOn Correct
+        [i|
+           f : (x : { v : Int | v >= 0 }) -> { v : Int | v > x }
            h : (x : { v : Int | v >= 0 }) -> { v : Int | v > x }
            g : (x : Int) -> (y : Int) -> (x' : { v : Int | v >= x }) -> (y' : { v : Int | v >= y }) -> { v : Int | v > x & v > y }
 

### 2020-06-29 toy/src/Toy/Language/EnvironmentUtils.hs
splitTypes only needs the type of the function, not a full FunSig

@@ -5,6 +5,7 @@ module Toy.Language.EnvironmentUtils
 , buildCombinedMapping
 , buildTypesMapping
 , retType
+, splitTypes
 
 , ArgTypes
 , Var2Ty
@@ -24,17 +25,15 @@ type Var2Ty = HM.HashMap VarName Ty
 funTypesMapping :: FunSig -> FunDefT a -> (ArgTypes, RefinedBaseTy)
 funTypesMapping sig def = (arg2type, resType)
   where
-    (argTypes, resType) = splitTypes sig
+    (argTypes, resType) = splitTypes $ funTy sig
     arg2type = zip (funArgs def) argTypes
 
 retType :: FunSig -> RefinedBaseTy
-retType = snd . splitTypes
+retType = snd . splitTypes . funTy
 
-splitTypes :: FunSig -> ([Ty], RefinedBaseTy)
-splitTypes = go . funTy
-  where
-    go (TyBase rbTy) = ([], rbTy)
-    go (TyArrow ArrowTy { .. }) = first (domTy :) $ go codTy
+splitTypes :: Ty -> ([Ty], RefinedBaseTy)
+splitTypes (TyBase rbTy) = ([], rbTy)
+splitTypes (TyArrow ArrowTy { .. }) = first (domTy :) $ splitTypes codTy
 
 buildVarsMap :: Monad m => (VarName -> Ty -> m a) -> ArgTypes -> m (HM.HashMap VarName a)
 buildVarsMap f args = HM.fromList <$> mapM sequence [ (var, f var ty) | (var, ty) <- args ]

### 2020-07-02 tex/paper.tex
Start playing around with ott/ottalt
More syntax defined
Try to fit the syntax in two subfigures
Start working on defining typing rules
Finally got the T-App rule to typecheck
ott is fun!
Add a refinement variable metavariable
Adding arrows is trivial now
WFT → TFW as a nicer abbreviation
Add lambda abstraction typing rule
Stub conditionals
Start working on the subtyping rules
Get rid of the binops for the core language
Add the TWF-BaseRef rule
Add the T-Sub rule
Rename BaseRef → TrueRef
Add TWF-Conj rule
Move \top to the term level, add \bot and add typing rules
Disambiguate surface language

@@
-\documentclass[12pt,a4paper]{article}
+\documentclass[a4paper]{article}
+\usepackage{a4wide}
 \usepackage[utf8]{inputenc}
 \usepackage[T2A]{fontenc}
+\usepackage{amsmath,amsfonts}
+\usepackage{xcolor}
+\usepackage{ottalt}
+\usepackage{subcaption}
+
+\addtolength{\topmargin}{-1.75in}
+\addtolength{\textheight}{2.75in}
+
+\begin{document}
+
+\newNTclass{nonterm}
+\newnonterm e \varepsilon
+\newnonterm{eS}{\varepsilon}
+\newnonterm{ec}{\overline\varepsilon}
+\newnonterm r \rho
+\newnonterm{tS}{\tau}
+\newnonterm{tC}{\overline\tau}
+\newnonterm s s
+\newnonterm B B
+\newnonterm{G}{\Gamma}
+\newnonterm{GC}{\overline\Gamma}
+\newnonterm{binop}{\odot}
+
+\newNTclass{gterm}
+\newgterm x x
+\newgterm v v
+\newgterm n n
+
+\inputott{surface.ott}
 
 \title{Compiling refinement types to dependent types}
-
-\begin{document}
 
@@
 
-\section{Calculus definition}
+\section{Calculi definitions}
 
@@
 
+\begin{figure}[ht]
+  \footnotesize
+  \begin{subfigure}{.6\textwidth}
+	\nonterms{eS}
+	\caption{Term level}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+	\nonterms{r,B,tS}
+	\caption{Type level}
+  \end{subfigure}
+  \caption{Surface language syntax}
+  \label{fig:surface_syntax}
+\end{figure}
+
+\begin{figure}[ht]
+  \footnotesize
+  \drules[TWF]{$\nonterm G \vdash \nonterm{tS} \text{ ok}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
+  \drules[T]{$\nonterm G \vdash \nonterm{eS} : \nonterm{tS}$}{term typing}{BoolTop,BoolBot,Var,Abs,App,If,Sub}
+  \drules[ST]{$\nonterm G \vdash \nonterm{tS}_1 <: \nonterm{tS}_2$}{subtyping}{Base,Arr}
+  \caption{Surface language typing rules}
+  \label{fig:surface_typing}
+\end{figure}
+
 \subsection{Core language}
+
+\begin{figure}[ht]
+  \footnotesize
+  \nonterms{ec,s}
+  \caption{Core language syntax}
+  \label{fig:core_syntax}
+\end{figure}
+
+\begin{figure}[ht]
+  \footnotesize
+  \drules[CT]{$\nonterm G \vdash \nonterm{ec}_1 : \nonterm{ec}_2$}{core typing}{Sort,Var,Weaken,Form,App,Abs,Conv}
+  \caption{Core language typing rules}
+  \label{fig:surface_typing}
+\end{figure}
 

### 2020-07-02 tex/build.sh
Make the build script error out on first failure

@@ -1,4 +1,6 @@
 #! /bin/sh
 
+set -e
+
 ott -tex_wrap false -i surface.ott -o surface.ott.tex
 pdflatex paper.tex

### 2020-07-02 tex/paper.tex
Clarify the syntax for typing contexts
Set up some derived entities in the core language

@@
 \newnonterm B B
-\newnonterm{G}{\Gamma}
+\newnonterm G \Gamma
 \newnonterm{GC}{\overline\Gamma}
@@
 
+We also establish some derived entities in the target language to simplify the subsequent exposition:
+
+\begin{itemize}
+  \item Non-dependent function type:
+	\[
+	  \tau_1 \rightarrow \tau_2 := \Pi x : \tau_1. \tau_2,
+	\]
+	where $\tau_2$ does not have $x$ free.
+  \item Dependent pair type, with the first component being a value $x$ of type $\tau$ and the second component being of type $P x$:
+	\[
+	  \{ x : \tau\ **\ P x \} := \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha.
+	\]
+  \item Accessing the first and second component of the dependent pair:
+	\[
+	  \text{fst} := TODO,
+	\]
+	\[
+	  \text{snd} := TODO.
+	\]
+  \item Non-dependent pair type:
+	\[
+	  (\tau_1, \tau_2) := \{ x : \tau_1\ **\ (\lambda \_ : \tau_1. \tau_2) x \}.
+	\]
+\end{itemize}
+
 \section{Translation}

### 2020-07-02 tex/paper.tex
Start writing about the translation function

@@
 
+The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic.
+In the subsequent text we assume the oracle can decide the resulting proposition
+(otherwise it might be left as a separate lemma to be filled out by the user in the core language, or rejected altogether).
+
+For the example oracle being an SMT solver and the syntax of terms also including mathematics, $[|\ |]$ might be defined as
+\begin{equation}
+\begin{aligned}
+  & [| \emptyset |] = \top																						\\
+  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] = [| \nonterm G |] \land [| \{x/v\} \nonterm r |]   \\
+  & [| \nonterm r_1 \land \nonterm r_2 |] = [| \nonterm r_1 |] \land [| \nonterm r_2 |]							\\
+  & \cdots TODO
+\end{aligned}
+\end{equation}
+
 \subsection{Core language}

### 2020-07-02 tex/paper.tex
Fix spaces
Use coloneqq
Describe terms translation

@@
 \usepackage{amsmath,amsfonts}
+\usepackage{mathtools}
 \usepackage{xcolor}
@@
 \begin{aligned}
-  & [| \emptyset |] = \top																						\\
-  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] = [| \nonterm G |] \land [| \{x/v\} \nonterm r |]   \\
-  & [| \nonterm r_1 \land \nonterm r_2 |] = [| \nonterm r_1 |] \land [| \nonterm r_2 |]							\\
+  & [| \emptyset |]                                         && \coloneqq \top																						 \\
+  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \coloneqq [| \nonterm G |] \land [| \{x/v\} \nonterm r |] \\
+  & [| \nonterm r_1 \land \nonterm r_2 |]                   && \coloneqq [| \nonterm r_1 |] \land [| \nonterm r_2 |]		 \\
   & \cdots TODO
@@
 
+%TODO define substitutions on the \nonterm r
+
 \subsection{Core language}
+
+We take $\lambda C$ \cite{Nederpelt14} enriched with integers and case-splitting as our core dependently typed language.
 
@@
   \[
-	  \tau_1 \rightarrow \tau_2 := \Pi x : \tau_1. \tau_2,
+    \tau_1 \rightarrow \tau_2 \coloneqq \Pi x : \tau_1. \tau_2,
   \]
   where $\tau_2$ does not have $x$ free.
-  \item Dependent pair type, with the first component being a value $x$ of type $\tau$ and the second component being of type $P x$:
+  \item Dependent pair type, or $\Sigma$-type, with the first component being a value $x$ of type $\tau$ and the second component being of type $P x$:
   \[
-	  \{ x : \tau\ **\ P x \} := \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha.
+    \{ x : \tau\ **\ P x \} \coloneqq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha.
   \]
@@
   \[
-	  \text{fst} := TODO,
+    \text{fst} \coloneqq TODO,
   \]
   \[
-	  \text{snd} := TODO.
+    \text{snd} \coloneqq TODO.
   \]
@@
   \[
-	  (\tau_1, \tau_2) := \{ x : \tau_1\ **\ (\lambda \_ : \tau_1. \tau_2) x \}.
+    (\tau_1, \tau_2) \coloneqq \{ x : \tau_1\ **\ (\lambda \_ : \tau_1. \tau_2) x \}.
   \]
@@
 
+We translate the surface language types and terms using the $\mu_\tau$ and $\mu_\varepsilon$ functions respectively.
+$\mu_\tau$ is defined as follows:
+\begin{equation}
+\begin{aligned}
+  & \mu_\tau( (x : \nonterm{tS}_1) \rightarrow \nonterm{tS}_2)  && \coloneqq \Pi x : \mu_\tau(\nonterm{tS}_1). \mu_\tau(\nonterm{tS}_2)  \\
+  & \mu_\tau( \{ v : \nonterm{tS} | \nonterm{r} \})             && \coloneqq \{ v : \mu_\tau(\nonterm{tS})\ **\ \mu_\tau(\nonterm r) \}  \\
+  & \mu_\tau( \nonterm r_1 \land \nonterm r_2 )                 && \coloneqq ( \mu_\tau(\nonterm r_1) , \mu_\tau(\nonterm r_2))          \\
+  & \mu_\tau( \nonterm{eS} )                                    && \coloneqq \mu_\varepsilon(\nonterm{eS}) = \text{True}                 \\
+  & \mu_\tau(B)                                                 && \coloneqq B                                                           \\
+\end{aligned}
+\end{equation}
+Or, in words:
+\begin{itemize}
+  \item (dependent) arrow types are translated to the corresponding $\Pi$-types,
+  \item refined types are translated to the corresponding $\Sigma$-types,
+  \item atomic refinements (which are $\text{Bool}$-typed terms) are translated to propositions stating that the corresponding term is equal to $\text{True}$,
+  \item conjunctions of several (atomic) refinements are represented as tuples of the corresponding atomic translations.
+\end{itemize}
+
 \end{document}

### 2020-07-03 tex/paper.tex
Fix duplicate labels
Use \triangleq instead of \coloneqq

@@
 \usepackage[T2A]{fontenc}
-\usepackage{amsmath,amsfonts}
+\usepackage{amsmath,amsfonts,amssymb}
 \usepackage{mathtools}
@@
 \begin{aligned}
-  & [| \emptyset |]                                         && \coloneqq \top																						 \\
-  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \coloneqq [| \nonterm G |] \land [| \{x/v\} \nonterm r |] \\
-  & [| \nonterm r_1 \land \nonterm r_2 |]                   && \coloneqq [| \nonterm r_1 |] \land [| \nonterm r_2 |]		 \\
+  & [| \emptyset |]                                         && \triangleq \top																						 \\
+  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \triangleq [| \nonterm G |] \land [| \{x/v\} \nonterm r |] \\
+  & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq [| \nonterm r_1 |] \land [| \nonterm r_2 |]		 \\
   & \cdots TODO
@@
   \caption{Core language typing rules}
-  \label{fig:surface_typing}
+  \label{fig:core_typing}
 \end{figure}
@@
   \[
-    \tau_1 \rightarrow \tau_2 \coloneqq \Pi x : \tau_1. \tau_2,
+    \tau_1 \rightarrow \tau_2 \triangleq \Pi x : \tau_1. \tau_2,
   \]
@@
   \[
-    \{ x : \tau\ **\ P x \} \coloneqq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha.
+    \{ x : \tau\ **\ P x \} \triangleq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha.
   \]
@@
   \[
-    \text{fst} \coloneqq TODO,
+    \text{fst} \triangleq TODO,
   \]
   \[
-    \text{snd} \coloneqq TODO.
+    \text{snd} \triangleq TODO.
   \]
@@
   \[
-    (\tau_1, \tau_2) \coloneqq \{ x : \tau_1\ **\ (\lambda \_ : \tau_1. \tau_2) x \}.
+    (\tau_1, \tau_2) \triangleq \{ x : \tau_1\ **\ (\lambda \_ : \tau_1. \tau_2) x \}.
   \]
@@
 \begin{aligned}
-  & \mu_\tau( (x : \nonterm{tS}_1) \rightarrow \nonterm{tS}_2)  && \coloneqq \Pi x : \mu_\tau(\nonterm{tS}_1). \mu_\tau(\nonterm{tS}_2)  \\
-  & \mu_\tau( \{ v : \nonterm{tS} | \nonterm{r} \})             && \coloneqq \{ v : \mu_\tau(\nonterm{tS})\ **\ \mu_\tau(\nonterm r) \}  \\
-  & \mu_\tau( \nonterm r_1 \land \nonterm r_2 )                 && \coloneqq ( \mu_\tau(\nonterm r_1) , \mu_\tau(\nonterm r_2))          \\
-  & \mu_\tau( \nonterm{eS} )                                    && \coloneqq \mu_\varepsilon(\nonterm{eS}) = \text{True}                 \\
-  & \mu_\tau(B)                                                 && \coloneqq B                                                           \\
+  & \mu_\tau( (x : \nonterm{tS}_1) \rightarrow \nonterm{tS}_2)  && \triangleq \Pi x : \mu_\tau(\nonterm{tS}_1). \mu_\tau(\nonterm{tS}_2)  \\
+  & \mu_\tau( \{ v : \nonterm{tS} | \nonterm{r} \})             && \triangleq \{ v : \mu_\tau(\nonterm{tS})\ **\ \mu_\tau(\nonterm r) \}  \\
+  & \mu_\tau( \nonterm r_1 \land \nonterm r_2 )                 && \triangleq ( \mu_\tau(\nonterm r_1) , \mu_\tau(\nonterm r_2))          \\
+  & \mu_\tau( \nonterm{eS} )                                    && \triangleq \mu_\varepsilon(\nonterm{eS}) = \text{True}                 \\
+  & \mu_\tau(B)                                                 && \triangleq B                                                           \\
 \end{aligned}

### 2020-07-05 tex/paper.tex
More consistent usage of the underscore to mark unused vars
Add a missing \cite
Somewhat better layout
More consistent usage of "target"/"core" language
tC isn't used
Core language doesn't have a terms/types distinction anyway at syntax
level.
More consistent terms casing

@@
 \usepackage{subcaption}
+\usepackage{cite}
 
-\addtolength{\topmargin}{-1.75in}
-\addtolength{\textheight}{2.75in}
+\addtolength{\topmargin}{-1.05in}
+\addtolength{\textheight}{1.55in}
 
@@
 \newnonterm e \varepsilon
-\newnonterm{eS}{\varepsilon}
+\newnonterm{es}{\varepsilon}
 \newnonterm{ec}{\overline\varepsilon}
 \newnonterm r \rho
-\newnonterm{tS}{\tau}
-\newnonterm{tC}{\overline\tau}
+\newnonterm{ts}{\tau}
 \newnonterm s s
@@
   \begin{subfigure}{.6\textwidth}
-	\nonterms{eS}
+	\nonterms{es}
 	\caption{Term level}
@@
   \begin{subfigure}{.5\textwidth}
-	\nonterms{r,B,tS}
+	\nonterms{r,B,ts}
 	\caption{Type level}
@@
   \footnotesize
-  \drules[TWF]{$\nonterm G \vdash \nonterm{tS} \text{ ok}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
-  \drules[T]{$\nonterm G \vdash \nonterm{eS} : \nonterm{tS}$}{term typing}{BoolTop,BoolBot,Var,Abs,App,If,Sub}
-  \drules[ST]{$\nonterm G \vdash \nonterm{tS}_1 <: \nonterm{tS}_2$}{subtyping}{Base,Arr}
+  \drules[TWF]{$\nonterm G \vdash \nonterm{ts} \text{ ok}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
+  \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{BoolTop,BoolBot,Var,Abs,App,If,Sub}
+  \drules[ST]{$\nonterm G \vdash \nonterm{ts}_1 <: \nonterm{ts}_2$}{subtyping}{Base,Arr}
   \caption{Surface language typing rules}
@@
 
-We also establish some derived entities in the target language to simplify the subsequent exposition:
+We also establish some derived entities in the core language to simplify the subsequent exposition:
 
@@
   \[
-    \tau_1 \rightarrow \tau_2 \triangleq \Pi x : \tau_1. \tau_2,
+    \tau_1 \rightarrow \tau_2 \triangleq \Pi \_ : \tau_1. \tau_2.
   \]
-  where $\tau_2$ does not have $x$ free.
   \item Dependent pair type, or $\Sigma$-type, with the first component being a value $x$ of type $\tau$ and the second component being of type $P x$:
@@
 \begin{aligned}
-  & \mu_\tau( (x : \nonterm{tS}_1) \rightarrow \nonterm{tS}_2)  && \triangleq \Pi x : \mu_\tau(\nonterm{tS}_1). \mu_\tau(\nonterm{tS}_2)  \\
-  & \mu_\tau( \{ v : \nonterm{tS} | \nonterm{r} \})             && \triangleq \{ v : \mu_\tau(\nonterm{tS})\ **\ \mu_\tau(\nonterm r) \}  \\
+  & \mu_\tau( (x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2)  && \triangleq \Pi x : \mu_\tau(\nonterm{ts}_1). \mu_\tau(\nonterm{ts}_2)  \\
+  & \mu_\tau( \{ v : \nonterm{ts} | \nonterm{r} \})             && \triangleq \{ v : \mu_\tau(\nonterm{ts})\ **\ \mu_\tau(\nonterm r) \}  \\
   & \mu_\tau( \nonterm r_1 \land \nonterm r_2 )                 && \triangleq ( \mu_\tau(\nonterm r_1) , \mu_\tau(\nonterm r_2))          \\
-  & \mu_\tau( \nonterm{eS} )                                    && \triangleq \mu_\varepsilon(\nonterm{eS}) = \text{True}                 \\
+  & \mu_\tau( \nonterm{es} )                                    && \triangleq \mu_\varepsilon(\nonterm{es}) = \text{True}                 \\
   & \mu_\tau(B)                                                 && \triangleq B                                                           \\

### 2020-07-05 tex/paper.tex
Theorem command definition
Write a bit more about oracle translation assumptions

@@
 \addtolength{\textheight}{1.55in}
+
+\newtheorem{theorem}{Theorem}
 
@@
 
-The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic.
-In the subsequent text we assume the oracle can decide the resulting proposition
-(otherwise it might be left as a separate lemma to be filled out by the user in the core language, or rejected altogether).
-
-For the example oracle being an SMT solver and the syntax of terms also including mathematics, $[|\ |]$ might be defined as
+The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic:
 \begin{equation}
@@
   & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq [| \nonterm r_1 |] \land [| \nonterm r_2 |]		  \\
-  & \cdots TODO
+  & [| \nonterm{es} |]                                      && \triangleq \text{oracle}(\nonterm{es})                     \\
+  & [| \nonterm{es}_1 = \nonterm{es}_2 |]                   && \triangleq \text{oracle}(\nonterm{es}_1 = \nonterm{es}_2)  \\
 \end{aligned}
@@
 
+The last two equations represent the assumption
+that the oracle knows how to translate the elementary refinements being the terms $\nonterm{es}$
+or equality propositions $\nonterm{es}_1 = \nonterm{es}_2$.
+
 %TODO define substitutions on the \nonterm r
+
+In the subsequent text we also assume the oracle can decide the resulting proposition
+(otherwise it might be left as a separate lemma to be filled out by the user in the core language, or rejected altogether).
 

### 2020-07-05 tex/paper.tex
A bit nicer equations display

@@
 \begin{aligned}
-  & [| \emptyset |]                                         && \triangleq \top																						\\
-  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \triangleq [| \nonterm G |] \land [| \{x/v\} \nonterm r |] \\
-  & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq [| \nonterm r_1 |] \land [| \nonterm r_2 |]		  \\
-  & [| \nonterm{es} |]                                      && \triangleq \text{oracle}(\nonterm{es})                     \\
-  & [| \nonterm{es}_1 = \nonterm{es}_2 |]                   && \triangleq \text{oracle}(\nonterm{es}_1 = \nonterm{es}_2)  \\
+  & [| \emptyset |]                                         && \triangleq && \top																						  \\
+  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \triangleq && [| \nonterm G |] \land [| \{x/v\} \nonterm r |]  \\
+  & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq && [| \nonterm r_1 |] \land [| \nonterm r_2 |]      \\
+  & [| \nonterm{es} |]                                      && \triangleq && \text{oracle}(\nonterm{es})                      \\
+  & [| \nonterm{es}_1 = \nonterm{es}_2 |]                   && \triangleq && \text{oracle}(\nonterm{es}_1 = \nonterm{es}_2)   \\
 \end{aligned}

### 2020-07-05 tex/paper.tex
State the theorem
Add some TODOs to ensure I don't forget about them

@@
 \section{Overview}
+
+%TODO describe the notion of the oracle (like an SMT solver) and what we expect of the oracle.
 
@@
   & \mu_\tau( \nonterm{es} )                                    && \triangleq \mu_\varepsilon(\nonterm{es}) = \text{True}                 \\
+%TODO add syntax/typing rules for True
   & \mu_\tau(B)                                                 && \triangleq B                                                           \\
@@
 
+%TODO define translation for gammas
+
+\subsection{Correctness}
+
+We next formulate our key theorem:
+
+\theorem For any well-typed surface language term $\nonterm{es}$, the result of the translation $\mu_\epsilon(\nonterm{es})$ is well-typed in the core language:
+\[
+  \nonterm G \vdash \nonterm{es} : \nonterm{ts} \implies \mu_\Gamma(\nonterm G) \vdash \mu_\epsilon(\nonterm{es}) : \mu_\tau(\nonterm{ts})
+\]
+
 \end{document}

### 2020-07-06 tex/paper.tex
Ok, I completely forgot the correct way of doing theorems

@@
 
-\theorem For any well-typed surface language term $\nonterm{es}$, the result of the translation $\mu_\epsilon(\nonterm{es})$ is well-typed in the core language:
+\begin{theorem}
+  For any well-typed surface language term $\nonterm{es}$,
+  the result of the translation $\mu_\epsilon(\nonterm{es})$ is well-typed in the core language:
   \[
@@
   \]
+\end{theorem}
 

### 2020-07-06 tex/paper.tex
Something weird with spacing there
We'll need equality as well
Using \overline for the core language is a really bad idea
We'll probably use it later to denote a sequence of terms/types in the
case-split rule.
Show the syntax for the contexts as well
Split surface language typing and subtyping
Something funky with spaces again
We actually have two key theorems

@@
 \newnonterm{es}{\varepsilon}
-\newnonterm{ec}{\overline\varepsilon}
+\newnonterm{ec}{\hat\varepsilon}
 \newnonterm r \rho
@@
 \newnonterm G \Gamma
-\newnonterm{GC}{\overline\Gamma}
+\newnonterm{GC}{\hat\Gamma}
 \newnonterm{binop}{\odot}
@@
   \begin{subfigure}{.5\textwidth}
-	\nonterms{r,B,ts}
+    \nonterms{G,r,B,ts}
     \caption{Type level}
@@
   \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{BoolTop,BoolBot,Var,Abs,App,If,Sub}
-  \drules[ST]{$\nonterm G \vdash \nonterm{ts}_1 <: \nonterm{ts}_2$}{subtyping}{Base,Arr}
-  \caption{Surface language typing rules}
+  \caption{Surface language typing}
   \label{fig:surface_typing}
@@
 
+\begin{figure}[ht]
+  \footnotesize
+  \drules[ST]{$\nonterm G \vdash \nonterm{ts}_1 <: \nonterm{ts}_2$}{subtyping}{Base,Arr}
+  \caption{Surface language subtyping}
+  \label{fig:surface_subtyping}
+\end{figure}
 The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic:
@@
     \]
+  \item Equality of expressions $\nonterm{ec}_1, \nonterm{ec}_2$ of type $S$:
+    \[
+      \nonterm{ec}_1 \equiv \nonterm{ec}_2 \triangleq \Pi P : S \rightarrow \star. (P x \rightarrow P y, P y \rightarrow P x).
+    \]
+    Careful reader will recognize this as the Leibniz equality \cite{FindSomethingForThis}.
 \end{itemize}
@@
 
-We next formulate our key theorem:
+The correctness of the translation hinges on two key theorems.
+The first one claims that a well-typed program in surface language remains well-typed after the translation to the core language.
+The second one claims that a well-typed surface language program evaluates to ``the same'' value as the translated one.
 
-\begin{theorem}
+\begin{theorem}[Well-typedness preservation]
   For any well-typed surface language term $\nonterm{es}$,
-  the result of the translation $\mu_\epsilon(\nonterm{es})$ is well-typed in the core language:
+  the result of the translation $\mu_\epsilon(\nonterm{es})$ is ``equivalently'' typeable in the core language:
   \[
@@
 \end{theorem}
+\begin{proof}
+\end{proof}
+
+\begin{theorem}[Evaluation equivalence]
+  For any well-typed surface language term $\nonterm{es}$, if $\nonterm{es}$ evaluates to v, then $\mu_\varepsilon(\nonterm{es})$ evaluates to $\mu_\varepsilon(v)$:
+  %TODO symbols and all that for v
+\end{theorem}
+\begin{proof}
+\end{proof}
 

### 2020-07-07 tex/paper.tex
Add a missing package
Add the stub context translation function
It doesn't make sense yet though.

@@
 \usepackage{cite}
+\usepackage{amsthm}
 
@@
 \end{figure}
+
 The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic:
@@
 
-We translate the surface language types and terms using the $\mu_\tau$ and $\mu_\varepsilon$ functions respectively.
+We translate the surface language contexts, types and terms into the corresponding core language entities
+using the $\mu_\Gamma$, $\mu_\tau$ and $\mu_\varepsilon$ functions respectively,
+which are defined mutually recursively.
+
+We start with defining $\mu_\Gamma$:
+\begin{equation}
+\begin{aligned}
+  & \mu_\Gamma(\emptyset)                                       && \triangleq \emptyset                                                     \\
+  & \mu_\Gamma(\nonterm G, \gterm x : \nonterm{ts})             && \triangleq \nonterm{GC},
+                                                                    \gterm x : \mu_\tau(\nonterm{ts})                                       \\
+  & \mu_\Gamma(\nonterm G, \nonterm{es}_1 = \nonterm{es}_2)     && \triangleq \nonterm{GC},
+                                                                    \mu_\varepsilon(\nonterm{es}_1) \equiv \mu_\varepsilon(\nonterm{es}_2)  \\
+\end{aligned}
+\end{equation}
+
 $\mu_\tau$ is defined as follows:
@@
 \end{itemize}
-
-%TODO define translation for gammas
 

### 2020-07-07 tex/paper.tex
Slightly better theorem wording

@@
 \begin{theorem}[Evaluation equivalence]
-  For any well-typed surface language term $\nonterm{es}$, if $\nonterm{es}$ evaluates to v, then $\mu_\varepsilon(\nonterm{es})$ evaluates to $\mu_\varepsilon(v)$:
+  If a well-typed surface language term $\nonterm{es}$ evaluates to v, then $\mu_\varepsilon(\nonterm{es})$ evaluates to $\mu_\varepsilon(v)$:
   %TODO symbols and all that for v

### 2020-07-07 tex/paper.tex
Remove term-only equation from the definition of [| |]

@@
   & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq && [| \nonterm r_1 |] \land [| \nonterm r_2 |]      \\
-  & [| \nonterm{es} |]                                      && \triangleq && \text{oracle}(\nonterm{es})                      \\
   & [| \nonterm{es}_1 = \nonterm{es}_2 |]                   && \triangleq && \text{oracle}(\nonterm{es}_1 = \nonterm{es}_2)   \\
@@
 
-The last two equations represent the assumption
-that the oracle knows how to translate the elementary refinements being the terms $\nonterm{es}$
-or equality propositions $\nonterm{es}_1 = \nonterm{es}_2$.
+The last equation represents the assumption
+that the oracle knows how to translate the atomic refinements.
 

### 2020-07-07 tex/paper.tex
Remove integers
No need to translate equality in contexts anymore
Also describe in words what \mu_\Gamma does

@@
 
-We take $\lambda C$ \cite{Nederpelt14} enriched with integers and case-splitting as our core dependently typed language.
+We take $\lambda C$ \cite{Nederpelt14} enriched with case-splitting as our core dependently typed language.
 
@@
 
-We start with defining $\mu_\Gamma$:
+We start with defining $\mu_\Gamma$,
+which merely translates the types of the bindings from the surface language into the core language:
 \begin{equation}
@@
                                                                     \gterm x : \mu_\tau(\nonterm{ts})                                       \\
-  & \mu_\Gamma(\nonterm G, \nonterm{es}_1 = \nonterm{es}_2)     && \triangleq \nonterm{GC},
-                                                                    \mu_\varepsilon(\nonterm{es}_1) \equiv \mu_\varepsilon(\nonterm{es}_2)  \\
 \end{aligned}

### 2020-07-07 tex/paper.tex
Proper recursion in the definition of \mu_\Gamma
Update the terms translation function

@@
 We start with defining $\mu_\Gamma$,
-which merely translates the types of the bindings from the surface language into the core language:
+which merely translates the types of the bindings in the context from the surface language into the core language:
 \begin{equation}
@@
   & \mu_\Gamma(\emptyset)                                       && \triangleq \emptyset                                                     \\
-  & \mu_\Gamma(\nonterm G, \gterm x : \nonterm{ts})             && \triangleq \nonterm{GC},
-                                                                    \gterm x : \mu_\tau(\nonterm{ts})                                       \\
+  & \mu_\Gamma(\nonterm G, \gterm x : \nonterm{ts})             && \triangleq \mu_\Gamma(\nonterm G), \gterm x : \mu_\tau(\nonterm{ts})     \\
 \end{aligned}
@@
   & \mu_\tau( \nonterm r_1 \land \nonterm r_2 )                 && \triangleq ( \mu_\tau(\nonterm r_1) , \mu_\tau(\nonterm r_2))                \\
-  & \mu_\tau( \nonterm{es} )                                    && \triangleq \mu_\varepsilon(\nonterm{es}) = \text{True}                 \\
+  & \mu_\tau( \nonterm{es}_1 = \nonterm{es}_2 )                 && \triangleq \mu_\varepsilon(\nonterm{es}_1) = \mu_\varepsilon(\nonterm{es}_2) \\
 %TODO add syntax/typing rules for True
@@
   \item refined types are translated to the corresponding $\Sigma$-types,
-  \item atomic refinements (which are $\text{Bool}$-typed terms) are translated to propositions stating that the corresponding term is equal to $\text{True}$,
+  \item atomic refinements, being propositions about surface language terms equality,
+    are translated to propositions stating the equality of translated core language terms,
   \item conjunctions of several (atomic) refinements are represented as tuples of the corresponding atomic translations.

### 2020-07-07 tex/paper.tex
Core language doesn't have any intrinsic equality

@@
   & \mu_\tau( \nonterm r_1 \land \nonterm r_2 )                 && \triangleq ( \mu_\tau(\nonterm r_1) , \mu_\tau(\nonterm r_2))                      \\
-  & \mu_\tau( \nonterm{es}_1 = \nonterm{es}_2 )                 && \triangleq \mu_\varepsilon(\nonterm{es}_1) = \mu_\varepsilon(\nonterm{es}_2) \\
+  & \mu_\tau( \nonterm{es}_1 = \nonterm{es}_2 )                 && \triangleq \mu_\varepsilon(\nonterm{es}_1) \equiv \mu_\varepsilon(\nonterm{es}_2)  \\
 %TODO add syntax/typing rules for True

### 2020-07-07 tex/paper.tex
Prepare reader for the slight notational abuse
Noticed one more case where the old-style subst is used

@@
   & [| \emptyset |]                                         && \triangleq && \top																						  \\
-  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \triangleq && [| \nonterm G |] \land [| \{x/v\} \nonterm r |]  \\
+  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \triangleq && [| \nonterm G |] \land [| [ v \mapsto x ] \nonterm r |]  \\
   & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq && [| \nonterm r_1 |] \land [| \nonterm r_2 |]      \\
@@
 
-$\mu_\tau$ is defined as follows:
+$\mu_\tau$ is defined as follows,
+with the slight abuse of notation to allow ``recursive'' calls
+on refinement part of the surface language syntax:
 \begin{equation}

### 2020-07-08 tex/paper.tex
Better description of the [| |] function
A bit more readability

@@
 
-The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic:
+The translation function $[|\ |]$ maps the typing context $\nonterm G$ to some logic
+that the oracle can handle.
+It might be defined as follows:
 \begin{equation}
 \begin{aligned}
-  & [| \emptyset |]                                         && \triangleq && \top																						  \\
-  & [| \nonterm G, x : \{ v : \nonterm B | \nonterm r \} |] && \triangleq && [| \nonterm G |] \land [| [ v \mapsto x ] \nonterm r |]  \\
-  & [| \nonterm r_1 \land \nonterm r_2 |]                   && \triangleq && [| \nonterm r_1 |] \land [| \nonterm r_2 |]      \\
-  & [| \nonterm{es}_1 = \nonterm{es}_2 |]                   && \triangleq && \text{oracle}(\nonterm{es}_1 = \nonterm{es}_2)   \\
+  & [|\ \emptyset\ |]                                         && \triangleq && \top																						            \\
+  & [|\ \nonterm G, x : \{ v : \nonterm B | \nonterm r \}\ |] && \triangleq && [|\ \nonterm G\ |] \land [|\ [ v \mapsto x ] \nonterm r\ |]  \\
+  & [|\ \nonterm r_1 \land \nonterm r_2\ |]                   && \triangleq && [|\ \nonterm r_1\ |] \land [|\ \nonterm r_2\ |]              \\
+  & [|\ \nonterm{es}_1 = \nonterm{es}_2\ |]                   && \triangleq && \text{oracle}(\nonterm{es}_1 = \nonterm{es}_2)             \\
 \end{aligned}

### 2020-07-08 tex/paper.tex
Update TODO

@@
   & \mu_\tau( \nonterm{es}_1 = \nonterm{es}_2 )                 && \triangleq \mu_\varepsilon(\nonterm{es}_1) \equiv \mu_\varepsilon(\nonterm{es}_2)  \\
-%TODO add syntax/typing rules for True
   & \mu_\tau(B)                                                 && \triangleq B                                                                       \\
+%TODO this last one is not well-defined
 \end{aligned}

### 2020-07-09 tex/paper.tex
Don't need a special bullet for the fst/snd functions
Sort of got the syntax for case
Add a note on the dummy parameters

@@
 \newnonterm s s
+\newnonterm l l
 \newnonterm B B
@@
   \begin{subfigure}{.6\textwidth}
-    \nonterms{es}
+    \nonterms{es,l}
     \caption{Term level}
@@
 \end{figure}
+
+The surface language might also have other base types (for example, $\mathbf{Int}$),
+elements of syntax (like numbers and operations on them)
+as well as typing rules relating those,
+but we omit them for the sake of brevity.
+
+Note that although the syntax does not allow refining function types or ADTs,
+it does not restrict the expressive power of the language.
+For instance, a refinement on a function that we might otherwise wish to write as
+\[
+  (x : \mathbf{Int}) \rightarrow \{ v : \mathbf{Int} \rightarrow \mathbf{Int} | v x = 0 \} \rightarrow \mathbf{Int}
+\]
+might instead be expressed via a dummy parameter:
+\[
+  (x : \mathbf{Int}) \rightarrow (f : \mathbf{Int} \rightarrow \mathbf{Int}) \rightarrow \{ \_ : \mathbf{Unit} | f x = 0 \} \rightarrow \mathbf{Int}
+\]
 
@@
     \[
-      \{ x : \tau\ **\ P x \} \triangleq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha.
+      \{ x : \tau\ **\ P x \} \triangleq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha,
     \]
-  \item Accessing the first and second component of the dependent pair:
+    with the corresponding accessors:
     \[

### 2020-07-09 tex/paper.tex
It's "projections", not "accessors"
There's gonna be a bunch of concessions

@@
 
-Note that although the syntax does not allow refining function types or ADTs,
-it does not restrict the expressive power of the language.
+We also have the following concessions:
+\begin{itemize}
+  \item The syntax only allows refinements on base types.
+    Note that this does not restrict the expressive power of the language.
     For instance, a refinement on a function that we might otherwise wish to write as
@@
     \[
-  (x : \mathbf{Int}) \rightarrow (f : \mathbf{Int} \rightarrow \mathbf{Int}) \rightarrow \{ \_ : \mathbf{Unit} | f x = 0 \} \rightarrow \mathbf{Int}
+      (x : \mathbf{Int}) \rightarrow (f : \mathbf{Int} \rightarrow \mathbf{Int}) \rightarrow \{ \_ : \mathbf{Unit} | f x = 0 \} \rightarrow \mathbf{Int}.
     \]
+\end{itemize}
 
@@
     \]
-    with the corresponding accessors:
+    with the corresponding projections:
     \[

### 2020-07-09 tex/paper.tex
Add shorthand \Int and \Unit commands

@@
 \newtheorem{theorem}{Theorem}
+
+\newcommand{\Unit}{\textbf{Unit}}
+\newcommand{\Int}{\textbf{Int}}
 
@@
 
-The surface language might also have other base types (for example, $\mathbf{Int}$),
+The surface language might also have other base types (for example, $\Int$),
 elements of syntax (like numbers and operations on them)
@@
     \[
-      (x : \mathbf{Int}) \rightarrow \{ v : \mathbf{Int} \rightarrow \mathbf{Int} | v x = 0 \} \rightarrow \mathbf{Int}
+      (x : \Int) \rightarrow \{ v : \Int \rightarrow \Int | v x = 0 \} \rightarrow \Int
     \]
@@
     \[
-      (x : \mathbf{Int}) \rightarrow (f : \mathbf{Int} \rightarrow \mathbf{Int}) \rightarrow \{ \_ : \mathbf{Unit} | f x = 0 \} \rightarrow \mathbf{Int}.
+      (x : \Int) \rightarrow (f : \Int \rightarrow \Int) \rightarrow \{ \_ : \Unit | f x = 0 \} \rightarrow \Int.
     \]

### 2020-07-09 tex/paper.tex
Add a note on ADTs having only a single field each
Remove bools
1. Units + ADTs are good enough to build up bools
2. We don't really need not use bools.
Write down what \top means now

@@
 
+\newcommand{\unit}{u}
 \newcommand{\Unit}{\textbf{Unit}}
@@
 
+We introduce a shorthand notation $\top$ denoting a tautologically true \emph{proposition},
+used for trivial refinements that do not carry any extra information.
+One good way to define $\top$ is the equality of the sole member of $\Unit$ to itself:
+$\top \triangleq \unit = \unit$.
+
 The surface language might also have other base types (for example, $\Int$),
@@
     \]
+  \item Our algebraic data types only allow a single ``field'' in each constructor.
+    One way to alleviate this is to introduce tuples,
+    but doing so would complicate the exposition
+    without any benefit for illustrating the main idea of this work.
 \end{itemize}

### 2020-07-10 tex/paper.tex
Don't forget to add references to case splitting
Add stub syntax for case analysis

@@
   \drules[TWF]{$\nonterm G \vdash \nonterm{ts} \text{ ok}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
-  \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{BoolTop,BoolBot,Var,Abs,App,If,Sub}
+  \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{Unit,Var,Abs,App,Case,Sub}
   \caption{Surface language typing}
@@
 
-We take $\lambda C$ \cite{Nederpelt14} enriched with case-splitting as our core dependently typed language.
+We take $\lambda C$ \cite{Nederpelt14} enriched with case-splitting inspired by \cite{TAPLVariants,EisenbergThesis} as our core dependently typed language.
 

### 2020-07-10 tex/paper.tex
Add T-Con rule for building ADTs
Minor wording fix

@@
   \drules[TWF]{$\nonterm G \vdash \nonterm{ts} \text{ ok}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
-  \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{Unit,Var,Abs,App,Case,Sub}
+  \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{Unit,Var,Abs,App,Case,Con,Sub}
   \caption{Surface language typing}
@@
 used for trivial refinements that do not carry any extra information.
-One good way to define $\top$ is the equality of the sole member of $\Unit$ to itself:
+One good way to define $\top$ is to state the equality of the sole inhabitant of $\Unit$ to itself:
 $\top \triangleq \unit = \unit$.
@@
     \]
-    Careful reader will recognize this as the Leibniz equality \cite{FindSomethingForThis}.
+    The reader might recognize this as the Leibniz equality \cite{FindSomethingForThis}.
 \end{itemize}

### 2020-07-10 tex/paper.tex
Add \tranctx, \tranty, \tranterm commands for consistency

@@
 
+\newcommand{\tranctx}{\mu_\Gamma}
+\newcommand{\tranty}{\mu_\tau}
+\newcommand{\tranterm}{\mu_\varepsilon}
+
 We translate the surface language contexts, types and terms into the corresponding core language entities
@@
 
-We start with defining $\mu_\Gamma$,
+We start with defining $\tranctx$,
 which merely translates the types of the bindings in the context from the surface language into the core language:
@@
 \begin{aligned}
-  & \mu_\Gamma(\emptyset)                                       && \triangleq \emptyset                                                     \\
-  & \mu_\Gamma(\nonterm G, \gterm x : \nonterm{ts})             && \triangleq \mu_\Gamma(\nonterm G), \gterm x : \mu_\tau(\nonterm{ts})     \\
+  & \tranctx(\emptyset)                                       && \triangleq \emptyset                                               \\
+  & \tranctx(\nonterm G, \gterm x : \nonterm{ts})             && \triangleq \tranctx(\nonterm G), \gterm x : \tranty(\nonterm{ts})  \\
 \end{aligned}
@@
 
-$\mu_\tau$ is defined as follows,
+$\tranty$ is defined as follows,
 with the slight abuse of notation to allow ``recursive'' calls
@@
 \begin{aligned}
-  & \mu_\tau( (x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2)  && \triangleq \Pi x : \mu_\tau(\nonterm{ts}_1). \mu_\tau(\nonterm{ts}_2)              \\
-  & \mu_\tau( \{ v : \nonterm{ts} | \nonterm{r} \})             && \triangleq \{ v : \mu_\tau(\nonterm{ts})\ **\ \mu_\tau(\nonterm r) \}              \\
-  & \mu_\tau( \nonterm r_1 \land \nonterm r_2 )                 && \triangleq ( \mu_\tau(\nonterm r_1) , \mu_\tau(\nonterm r_2))                      \\
-  & \mu_\tau( \nonterm{es}_1 = \nonterm{es}_2 )                 && \triangleq \mu_\varepsilon(\nonterm{es}_1) \equiv \mu_\varepsilon(\nonterm{es}_2)  \\
-  & \mu_\tau(B)                                                 && \triangleq B                                                                       \\
+  & \tranty( (x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2)  && \triangleq \Pi x : \tranty(\nonterm{ts}_1). \tranty(\nonterm{ts}_2)     \\
+  & \tranty( \{ v : \nonterm{ts} | \nonterm{r} \})             && \triangleq \{ v : \tranty(\nonterm{ts})\ **\ \tranty(\nonterm r) \}     \\
+  & \tranty( \nonterm r_1 \land \nonterm r_2 )                 && \triangleq ( \tranty(\nonterm r_1) , \tranty(\nonterm r_2))             \\
+  & \tranty( \nonterm{es}_1 = \nonterm{es}_2 )                 && \triangleq \tranterm(\nonterm{es}_1) \equiv \tranterm(\nonterm{es}_2)   \\
+  & \tranty(B)                                                 && \triangleq B                                                            \\
 %TODO this last one is not well-defined

### 2020-07-10 tex/paper.tex
Wrap translations for contexts and terms into paragraphs

@@
 
+\paragraph{Contexts.}
 We start with defining $\tranctx$,
@@
 
+\paragraph{Types.}
 $\tranty$ is defined as follows,
 with the slight abuse of notation to allow ``recursive'' calls
-on refinement part of the surface language syntax:
+on the refinement part of the surface language syntax:
 \begin{equation}

### 2020-07-11 tex/paper.tex
Add units to the core language

@@
 
-We take $\lambda C$ \cite{Nederpelt14} enriched with case-splitting inspired by \cite{TAPLVariants,EisenbergThesis} as our core dependently typed language.
+We take $\lambda C$ \cite{Nederpelt14}
+enriched with case-splitting inspired by \cite{TAPLVariants,EisenbergThesis}
+and a unit type as our core dependently typed language.
 
@@
   \footnotesize
-  \drules[CT]{$\nonterm G \vdash \nonterm{ec}_1 : \nonterm{ec}_2$}{core typing}{Sort,Var,Weaken,Form,App,Abs,Conv}
+  \drules[CT]{$\nonterm G \vdash \nonterm{ec}_1 : \nonterm{ec}_2$}{core typing}{Sort,UnitType,UnitTerm,Var,Weaken,Form,App,Abs,Conv}
   \caption{Core language typing rules}

### 2020-07-11 tex/paper.tex
Better wording for the \top notation

@@
 
-We introduce a shorthand notation $\top$ denoting a tautologically true \emph{proposition},
-used for trivial refinements that do not carry any extra information.
-One good way to define $\top$ is to state the equality of the sole inhabitant of $\Unit$ to itself:
+We introduce a shorthand notation $\top$ denoting some tautology
+which we use for trivial refinements that do not carry any extra information.
+In particular, one good way to define $\top$ is
+to state the equality of the sole inhabitant of $\Unit$ to itself:
 $\top \triangleq \unit = \unit$.

### 2020-07-11 tex/paper.tex
Actually add the commands for core units
Start working on the terms translation
Use \tran* things in the theorems as well

@@
 \newcommand{\Unit}{\textbf{Unit}}
+\newcommand{\unitc}{\hat{u}}
+\newcommand{\Unitc}{\widehat{\textbf{Unit}}}
 \newcommand{\Int}{\textbf{Int}}
@@
 
+\paragraph{Terms.}
+Finally we define $\tranterm$:
+\begin{equation}
+\begin{aligned}
+  & \tranterm(x) &&
+    \triangleq x                                                              \\
+  & \tranterm(\lambda x : \nonterm{ts}. \nonterm{es})  &&
+    \triangleq \lambda x : \tranty(\nonterm{ts}). \tranterm(\nonterm{es})     \\
+  & \tranterm(\nonterm{es}_1 \nonterm{es}_2)  &&
+    \triangleq \tranterm(\nonterm{es}_1) \tranterm(\nonterm{es}_2)            \\
+\end{aligned}
+\end{equation}
+
 \subsection{Correctness}
@@
   For any well-typed surface language term $\nonterm{es}$,
-  the result of the translation $\mu_\epsilon(\nonterm{es})$ is ``equivalently'' typeable in the core language:
+  the result of the translation $\tranterm(\nonterm{es})$ is ``equivalently'' typeable in the core language:
   \[
-    \nonterm G \vdash \nonterm{es} : \nonterm{ts} \implies \mu_\Gamma(\nonterm G) \vdash \mu_\epsilon(\nonterm{es}) : \mu_\tau(\nonterm{ts})
+    \nonterm G \vdash \nonterm{es} : \nonterm{ts} \implies \tranctx(\nonterm G) \vdash \tranterm(\nonterm{es}) : \tranty(\nonterm{ts})
   \]
@@
 \begin{theorem}[Evaluation equivalence]
-  If a well-typed surface language term $\nonterm{es}$ evaluates to v, then $\mu_\varepsilon(\nonterm{es})$ evaluates to $\mu_\varepsilon(v)$:
+  If a well-typed surface language term $\nonterm{es}$ evaluates to v, then $\tranterm(\nonterm{es})$ evaluates to $\tranterm(v)$:
   %TODO symbols and all that for v

### 2020-07-11 toy/src/Toy/Language/Solver/QueryInterp.hs
We'll also need Z3 functions represented

@@ -15,14 +15,18 @@ import Toy.Language.Solver.Types
 import Toy.Language.Syntax
 
 newtype Z3Var = Z3Var { getZ3Var :: AST } deriving (Show)
+newtype Z3Fun = Z3Fun { getZ3Fun :: FuncDecl } deriving (Show)
 
-newtype ConvertState = ConvertState { variables :: HM.HashMap VarName Z3Var } deriving (Show)
+data ConvertState = ConvertState
+  { variables :: HM.HashMap VarName Z3Var
+  , functions :: HM.HashMap VarName Z3Fun
+  } deriving (Show)
 
 type MonadConvert m = (MonadZ3 m, MonadState ConvertState m)
 
 solveDef :: FunSig -> QFunDef -> IO (SolveRes, VCFunDef SolveRes)
 solveDef funSig funDef = evalZ3 $ do
-  (intrAssertions, qTerm) <- evalStateT buildAsts $ ConvertState mempty
+  (intrAssertions, qTerm) <- evalStateT buildAsts $ ConvertState mempty mempty
   assert $ getIntrinsicAssertions intrAssertions
   sTerm <- solveQTerm qTerm
   pure (isAllCorrect $ query <$> sTerm, funDef { funBody = sTerm })

### 2020-07-11 toy/src/Toy/Language/Solver/QueryInterp.hs
Creating variables and creating functions are two different things

@@ -42,10 +42,10 @@ solveDef funSig funDef = evalZ3 $ do
       pure (intrAssertions, qTerm)
 
 initArgVars :: MonadConvert m => ArgTypes -> m ()
-initArgVars = mapM_ $ \(varName, ty) -> createVar ty varName
+initArgVars = mapM_ $ \(varName, ty) -> createName ty varName
 
 initRefVars :: MonadConvert m => RefAnnTerm -> m ()
-initRefVars = mapM_ $ \RefAnn { .. } -> createVar tyAnn $ subjectVar intrinsic
+initRefVars = mapM_ $ \RefAnn { .. } -> createName tyAnn $ subjectVar intrinsic
 
 newtype IntrinsicAssertions = IntrinsicAssertions { getIntrinsicAssertions :: AST }
 
@@ -59,7 +59,7 @@ convertQuery :: MonadConvert m => Query -> m AST
 convertQuery (q1 :& q2) = join $ (\a b -> mkAnd [a, b]) <$> convertQuery q1 <*> convertQuery q2
 convertQuery (antecedent :=> consequent) = do
   -- TODO find a better way
-  v' <- createVar (TyBase $ RefinedBaseTy TInt trueRefinement) $ subjectVar antecedent
+  v' <- createVar TInt $ subjectVar antecedent
   antecedent' <- convertRefinement antecedent
   consequent' <- convertRefinement consequent
   implication <- mkImplies antecedent' consequent'
@@ -104,20 +104,24 @@ convertRefinement Refinement { .. } = mapM (convertTerm . getARTerm) conjuncts >
       BinOpGt -> mkGt
       BinOpGeq -> mkGe
 
-createVar :: MonadConvert m => Ty -> VarName -> m Z3Var
-createVar ty varName = do
+createName :: MonadConvert m => Ty -> VarName -> m ()
+createName (TyBase rbty) varName = void $ createVar (baseType rbty) varName
+createName TyArrow {} varName = mkStringSymbol name >>= mkUninterpretedSort >>= void . mkFreshConst name
+  where
+    name = getName varName
+
+createVar :: MonadConvert m => BaseTy -> VarName -> m Z3Var
+createVar rbty varName = do
   ifM (gets $ HM.member varName . variables)
     (error [i|#{getName varName} has already been instantiated|])
     (pure ())
-  z3var <- Z3Var <$> mkFreshTypedVar ty (getName varName)
+  z3var <- Z3Var <$> mkFreshTypedVar rbty (getName varName)
[…]

### 2020-07-11 toy/src/Toy/Language/Solver/QueryInterp.hs
Better function creation
Interpret them as functions as opposed to uninterpreted sorts.

@@ -106,9 +106,7 @@ convertRefinement Refinement { .. } = mapM (convertTerm . getARTerm) conjuncts >
 
 createName :: MonadConvert m => Ty -> VarName -> m ()
 createName (TyBase rbty) varName = void $ createVar (baseType rbty) varName
-createName TyArrow {} varName = mkStringSymbol name >>= mkUninterpretedSort >>= void . mkFreshConst name
-  where
-    name = getName varName
+createName (TyArrow arrTy) varName = createFun arrTy varName
 
 createVar :: MonadConvert m => BaseTy -> VarName -> m Z3Var
 createVar rbty varName = do
@@ -122,6 +120,20 @@ createVar rbty varName = do
     mkFreshTypedVar = \case TInt -> mkFreshIntVar
                             TBool -> mkFreshBoolVar
                             TIntList -> error "TODO TIntList unsupported" -- TODO
+
+createFun :: MonadConvert m => ArrowTy -> VarName -> m ()
+createFun arrTy varName = do
+  args' <- mapM mkSort args
+  ret' <- mkSort (TyBase ret)
+  z3fun <- Z3Fun <$> mkFreshFuncDecl (getName varName) args' ret'
+  modify' $ \cs -> cs { functions = HM.insert varName z3fun $ functions cs }
+  where
+    (args, ret) = splitTypes $ TyArrow arrTy
+    mkSort (TyBase RefinedBaseTy { .. }) = case baseType of
+                                                TInt -> mkIntSort
+                                                TBool -> mkBoolSort
+                                                TIntList -> error "TODO TIntList unsupported"
+    mkSort _ = undefined -- TODO mkUninterpretedSort
 
 getVar :: MonadConvert m => VarName -> m Z3Var
 getVar varName = gets $ (HM.! varName) . variables

### 2020-07-11 toy/src/Toy/Language/Solver/QueryInterp.hs
More uniform casing

@@ -105,15 +105,15 @@ convertRefinement Refinement { .. } = mapM (convertTerm . getARTerm) conjuncts >
       BinOpGeq -> mkGe
 
 createName :: MonadConvert m => Ty -> VarName -> m ()
-createName (TyBase rbty) varName = void $ createVar (baseType rbty) varName
+createName (TyBase rbTy) varName = void $ createVar (baseType rbTy) varName
 createName (TyArrow arrTy) varName = createFun arrTy varName
 
 createVar :: MonadConvert m => BaseTy -> VarName -> m Z3Var
-createVar rbty varName = do
+createVar rbTy varName = do
   ifM (gets $ HM.member varName . variables)
     (error [i|#{getName varName} has already been instantiated|])
     (pure ())
-  z3var <- Z3Var <$> mkFreshTypedVar rbty (getName varName)
+  z3var <- Z3Var <$> mkFreshTypedVar rbTy (getName varName)
   modify' $ \cs -> cs { variables = HM.insert varName z3var $ variables cs }
   pure z3var
   where

### 2020-07-12 toy/src/Toy/Language/Solver/QueryInterp.hs
Eta reduce

@@ -105,8 +105,8 @@ convertRefinement Refinement { .. } = mapM (convertTerm . getARTerm) conjuncts >
       BinOpGeq -> mkGe
 
 createName :: MonadConvert m => Ty -> VarName -> m ()
-createName (TyBase rbTy) varName = void $ createVar (baseType rbTy) varName
-createName (TyArrow arrTy) varName = createFun arrTy varName
+createName (TyBase rbTy) = void . createVar (baseType rbTy)
+createName (TyArrow arrTy) = createFun arrTy
 
 createVar :: MonadConvert m => BaseTy -> VarName -> m Z3Var
 createVar rbTy varName = do

### 2020-07-12 toy/src/Toy/Language/Solver/QueryInterp.hs
Factor out `checkExists` and check functions don't exist

@@ -108,11 +108,14 @@ createName :: MonadConvert m => Ty -> VarName -> m ()
 createName (TyBase rbTy) = void . createVar (baseType rbTy)
 createName (TyArrow arrTy) = createFun arrTy
 
+checkExists :: MonadState s m => VarName -> (s -> HM.HashMap VarName v) -> m ()
+checkExists varName getter = ifM (gets $ HM.member varName . getter)
+                                 (error [i|#{getName varName} has already been instantiated|])
+                                 (pure ())
+
 createVar :: MonadConvert m => BaseTy -> VarName -> m Z3Var
 createVar rbTy varName = do
-  ifM (gets $ HM.member varName . variables)
-    (error [i|#{getName varName} has already been instantiated|])
-    (pure ())
+  checkExists varName variables
   z3var <- Z3Var <$> mkFreshTypedVar rbTy (getName varName)
   modify' $ \cs -> cs { variables = HM.insert varName z3var $ variables cs }
   pure z3var
@@ -123,6 +126,7 @@ createVar rbTy varName = do
 
 createFun :: MonadConvert m => ArrowTy -> VarName -> m ()
 createFun arrTy varName = do
+  checkExists varName functions
   args' <- mapM mkSort args
   ret' <- mkSort (TyBase ret)
   z3fun <- Z3Fun <$> mkFreshFuncDecl (getName varName) args' ret'

### 2020-07-12 toy/src/Toy/Language/Solver/QueryInterp.hs
Funs getter

@@ -142,6 +142,9 @@ createFun arrTy varName = do
 getVar :: MonadConvert m => VarName -> m Z3Var
 getVar varName = gets $ (HM.! varName) . variables
 
+getFun :: MonadConvert m => VarName -> m Z3Fun
+getFun varName = gets $ (HM.! varName) . functions
+
 mkAnd' :: (Foldable t, MonadZ3 m) => t AST -> m AST
 mkAnd' = mk . toList
   where

### 2020-07-12 toy/src/Toy/Language/Solver/QueryInterp.hs
Make `convertTerm` a top-level function

@@ -77,23 +77,25 @@ solveQTerm = onVCTerm $ \queryAST -> local $ do
 
 convertRefinement :: MonadConvert m => Refinement -> m AST
 convertRefinement Refinement { .. } = mapM (convertTerm . getARTerm) conjuncts >>= mkAnd'
+
+convertTerm :: MonadConvert m => Term -> m AST
+convertTerm = \case
+  TName _ varName -> getZ3Var <$> getVar varName
+  TInteger _ n -> mkIntNum n
+  TBinOp _ t1 op t2 -> join $ convertBinOp op <$> convertTerm t1 <*> convertTerm t2
+  TApp _ fun arg -> convertFunApp fun arg
+  TIfThenElse { .. } -> do
+    tthenCond' <- convertTerm tcond
+    tthen' <- convertTerm tthen
+
+    telseCond' <- mkNot tthenCond'
+    telse' <- convertTerm telse
+
+    thenBranch <- mkImplies tthenCond' tthen'
+    elseBranch <- mkImplies telseCond' telse'
+
+    mkAnd [thenBranch, elseBranch]
   where
-    convertTerm = \case
-      TName _ varName -> getZ3Var <$> getVar varName
-      TInteger _ n -> mkIntNum n
-      TBinOp _ t1 op t2 -> join $ convertBinOp op <$> convertTerm t1 <*> convertTerm t2
-      TApp {} -> error "fun app at refinement level unsupported yet" -- TODO
-      TIfThenElse { .. } -> do
-        tthenCond' <- convertTerm tcond
-        tthen' <- convertTerm tthen
-
-        telseCond' <- mkNot tthenCond'
-        telse' <- convertTerm telse
-
-        thenBranch <- mkImplies tthenCond' tthen'
-        elseBranch <- mkImplies telseCond' telse'
-
-        mkAnd [thenBranch, elseBranch]
     convertBinOp = \case
       BinOpPlus -> \a b -> mkAdd [a, b]
       BinOpMinus -> \a b -> mkAdd [a, b]
@@ -103,6 +105,9 @@ convertRefinement Refinement { .. } = mapM (convertTerm . getARTerm) conjuncts >
       BinOpNEq -> \a b -> mkEq a b >>= mkNot
       BinOpGt -> mkGt
       BinOpGeq -> mkGe
+
+convertFunApp :: MonadConvert m => Term -> Term -> m AST
+convertFunApp fun arg = undefined
 
[…]

### 2020-07-12 toy/src/Toy/Language/Solver/QueryInterp.hs
Stub fun app conversion

@@ -107,7 +107,14 @@ convertTerm = \case
       BinOpGeq -> mkGe
 
 convertFunApp :: MonadConvert m => Term -> Term -> m AST
-convertFunApp fun arg = undefined
+convertFunApp fun arg = go fun [arg]
+  where
+    go (TName _ varName) args = do
+      z3fun <- getFun varName
+      argASTs <- mapM convertTerm args
+      mkApp (getZ3Fun z3fun) argASTs
+    go (TApp _ fun' arg') args = go fun' (arg' : args)
+    go _ _ = error [i|Unsupported fun term type|]
 
 createName :: MonadConvert m => Ty -> VarName -> m ()
 createName (TyBase rbTy) = void . createVar (baseType rbTy)

### 2020-07-12 toy/src/Toy/Language/Solver/QueryInterp.hs
Factor out common accessors

@@ -152,10 +152,13 @@ createFun arrTy varName = do
     mkSort _ = undefined -- TODO mkUninterpretedSort
 
 getVar :: MonadConvert m => VarName -> m Z3Var
-getVar varName = gets $ (HM.! varName) . variables
+getVar = getNamed variables
 
 getFun :: MonadConvert m => VarName -> m Z3Fun
-getFun varName = gets $ (HM.! varName) . functions
+getFun = getNamed functions
+
+getNamed :: MonadState s m => (s -> HM.HashMap VarName v) -> VarName -> m v
+getNamed getter varName = gets $ (HM.! varName) . getter
 
 mkAnd' :: (Foldable t, MonadZ3 m) => t AST -> m AST
 mkAnd' = mk . toList

### 2020-07-14 tex/paper.tex
We indeed are going to prove WTP by induction on the typing derivation
Some stub text prior to the definition of the surface language

@@
 \subsection{Surface language}
+
+Our surface language is based on simply typed lambda calculus.
+Refinements over the types are, of course, the most important extension.
+The other ones are:
+\begin{itemize}
+  \item dependent arrow types, to allow subsequent refinements refer preceding arguments,
+  \item limited form of algebraic data types and dependent pattern matching,
+    to illustrate reasoning about those widely used constructs.
+\end{itemize}
 
@@
 \begin{proof}
+  We prove this by induction on the derivation of the typeability of $\nonterm{es}$.
+  In particular, we consider the last typing rule used:
 \end{proof}

### 2020-07-14 tex/build.sh
Start working on the bibliography

@@ -3,4 +3,4 @@
 set -e
 
 ott -tex_wrap false -i surface.ott -o surface.ott.tex
-pdflatex paper.tex
+pdflatex paper.tex && bibtex paper.aux && pdflatex paper.tex && pdflatex paper.tex

### 2020-07-15 tex/paper.tex
Start working on the bibliography
We probably won't have ADT/PM in our core language
Build fix
Add references for ADT/PM for the surface language

@@
 \usepackage{cite}
+\usepackage{url}
+\usepackage{babelbib}
+\usepackage[english]{babel}
 \usepackage{amsthm}
@@
   \item dependent arrow types, to allow subsequent refinements refer preceding arguments,
-  \item limited form of algebraic data types and dependent pattern matching,
+  \item limited form of algebraic data types and dependent pattern matching
+    insired by \cite{TAPLVariants,Eisenberg16},
     to illustrate reasoning about those widely used constructs.
@@
 
-We take $\lambda C$ \cite{Nederpelt14}
-enriched with case-splitting inspired by \cite{TAPLVariants,EisenbergThesis}
-and a unit type as our core dependently typed language.
+We take $\lambda C$ \cite{TTFPLambdaC}
+with a unit type as our core dependently typed language.
 
@@
 
+\bibliographystyle{babunsrt-lf}
+\bibliography{biblio}
+
 \end{document}

### 2020-07-16 tex/paper.tex
Stub BB encoding for ADTs
Write down what \tau means
Use parens instead of curlies for DPair
Better \usepackage grouping
Add the constructor for dpairs
Write DPair projections too
Stub BB encoding for ADTs in the core language

@@
 \usepackage[T2A]{fontenc}
-\usepackage{amsmath,amsfonts,amssymb}
+\usepackage{amsmath,amsfonts,amssymb,amsthm}
 \usepackage{mathtools}
@@
 \usepackage{subcaption}
-\usepackage{cite}
-\usepackage{url}
-\usepackage{babelbib}
+\usepackage{cite,url,babelbib}
 \usepackage[english]{babel}
-\usepackage{amsthm}
 
@@
 
+Although the syntax doesn't mention $\tau$, we will use it as a metavariable
+where a type (an inhabitant of $\star$) is expected.
+
+\newcommand{\dast}{\ **\ }
+
 We also establish some derived entities in the core language to simplify the subsequent exposition:
@@
     \[
-      \{ x : \tau\ **\ P x \} \triangleq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha,
+      (x : \tau \dast P x) \triangleq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha,
     \]
-    with the corresponding projections:
+    with the corresponding constructor (we overload the syntax here to resemble Idris):
     \[
-      \text{fst} \triangleq TODO,
+      (x \dast p) \triangleq \lambda \alpha : \star. \lambda f : (\Pi x' : \tau. P x' \rightarrow \alpha). f x p
+    \]
+    and projections
+    \[
+      \text{fst} \triangleq \lambda \sigma : (x : \tau \dast P x). \sigma \tau (\lambda x' : \tau. \lambda \_ : P x'. x'),
     \]
     \[
-      \text{snd} \triangleq TODO.
+      \text{snd} \triangleq \lambda \sigma : (x : \tau \dast P x). \sigma \tau (\lambda x' : \tau. \lambda p : P x'. p),
     \]
@@
     The reader might recognize this as the Leibniz equality \cite{FindSomethingForThis}.
+  \item ADTs, expressed using an encoding similar to Boehm-Berarducci's \cite{Bohm85}.
+    We borrow the syntax for the ADTs from the surface language and put a hat on it,
+    so let's take some ADT $\widehat{\langle \overline{l_i : \nonterm{ts}_i}^i \rangle}$.
+    The usual Boehm-Berarducci encoding associates the type
+    $\Pi \alpha : \star. \overline{(\nonterm{ts}_i \rightarrow \alpha)}^i \rightarrow \alpha$ with it.
+
+    We go a step further to account for the subset of the dependent pattern matching
+    that we are interested in.
+
+    First, each label $l_i$ gives rise to a term $\hat{l_i} : \nonterm{ts}_i \rightarrow \xi$.
+    %TODO operational semantics?
+    We then define the derived form
+    \[
+      \widehat{\langle \overline{l_i : \nonterm{ts}_i}^i \rangle}
+        \triangleq
+          \Pi s : \xi.
+          \Pi \alpha : \star.
+          \overline{(\Pi x_i : \nonterm{ts}_i. s \equiv \hat{l_i} x_i \rightarrow \alpha)}^i \rightarrow \alpha.
+    \]
+    The outer extra argument named $s$ denotes the scrutinee of the case expression to be defined next.
+    The inner extra argument with the type $s \equiv \hat{l_i} x_i$
+    expresses the subset of dependent pattern matching
+    that we rely on for the translation from the surface language:
+    in particular, it provides the ``case branch'' with a proof
+    that the scrutinee $s$ is indeed equal to the ``case pattern'' $\hat{l_i} x_i$.
 \end{itemize}

### 2020-07-18 tex/paper.tex
Minor stylistic fixes
More minor stylistic fix
Committing because I'm upset the derived form for dependent pattern
matching just doesn't come to mind :(

@@
 
-We also establish some derived entities in the core language to simplify the subsequent exposition:
+We also establish some derived forms in the core language to simplify the subsequent exposition:
 
@@
     \[
-      \text{snd} \triangleq \lambda \sigma : (x : \tau \dast P x). \sigma \tau (\lambda x' : \tau. \lambda p : P x'. p),
+      \text{snd} \triangleq \lambda \sigma : (x : \tau \dast P x). \sigma \tau (\lambda x' : \tau. \lambda p : P x'. p).
     \]
@@
 
-    First, each label $l_i$ gives rise to a term $\hat{l_i} : \nonterm{ts}_i \rightarrow \xi$.
+    First, each label $l_i$ gives rise to a term $\hat{l_i}$ of type $\nonterm{ts}_i \rightarrow \xi$.
     %TODO operational semantics?

### 2020-07-18 tex/paper.tex
More instances of old \mu_ things

@@
 We translate the surface language contexts, types and terms into the corresponding core language entities
-using the $\mu_\Gamma$, $\mu_\tau$ and $\mu_\varepsilon$ functions respectively,
+using the $\tranctx$, $\tranty$ and $\tranterm$ functions respectively,
 which are defined mutually recursively.

### 2020-07-18 tex/paper.tex
Introduce translating typing relations and explain why we need that

@@
 \newcommand{\tranty}{\mu_\tau}
+\newcommand{\tranderiv}{\mu_\vdash}
 \newcommand{\tranterm}{\mu_\varepsilon}
@@
 We translate the surface language contexts, types and terms into the corresponding core language entities
-using the $\tranctx$, $\tranty$ and $\tranterm$ functions respectively,
+using the $\tranctx$, $\tranty$ and $\tranderiv$ functions respectively,
 which are defined mutually recursively.
+
+Before we dive into defining these functions, $\tranderiv$ deserves a few introductory words.
+Although the aim is to translate terms,
+what this function really does is translating the surface typing derivation for a given (well-typed) term
+into a typing derivation in the core language.
+The reason for the extra complexity stems from the \ottdrulename{T-Sub} typing rule.
+In particular, this rule invokes the subtyping check, which, in turn, invokes the oracle.
+The oracle then might produce a core language term proving that the subtyping relation in question holds,
+and the presence of this rule is in some sense a carrier of this proof term.
 

### 2020-07-19 tex/paper.tex
Use \ranchor instead of \ottdrulename
Start defining term translation

@@
 into a typing derivation in the core language.
-The reason for the extra complexity stems from the \ottdrulename{T-Sub} typing rule.
+The reason for the extra complexity stems from the \ranchor{T-Sub} typing rule.
 In particular, this rule invokes the subtyping check, which, in turn, invokes the oracle.
@@
 \paragraph{Terms.}
-Finally we define $\tranterm$:
+Finally we define $\tranderiv$.
+As it is defined on typing derivation trees,
+we consider the term typing rule that is used last to prove $\nonterm G \vdash \nonterm{es} : \nonterm{ts}$.
+
+\begin{itemize}
+  \item \ranchor{T-Unit}.
+    This rule has no premises, so we just translate it into \ranchor{CT-UnitTerm}:
+    \[
+      \inferrule{ }{\nonterm G \vdash \unit : \{ \gterm v : \Unit | \top \}} \longmapsto \inferrule{ }{\nonterm{GC} \vdash \unitc : \Unitc}
+    \]
+\end{itemize}
+
 \begin{equation}

### 2020-07-19 tex/paper.tex
Add a conjecture that units don't break λC
No need for `ok` in `Γ |- τ ok`

@@
   \footnotesize
-  \drules[TWF]{$\nonterm G \vdash \nonterm{ts} \text{ ok}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
+  \drules[TWF]{$\nonterm G \vdash \nonterm{ts}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
   \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{Unit,Var,Abs,App,Case,Con,Sub}
@@
 with a unit type as our core dependently typed language.
+We conjecture that the introduction of units
+does not break the usual metatheoretical properties of the system.
 

### 2020-07-19 tex/paper.tex
Add context well-formedness rules
Unite typing and subtyping rules figures
Refer the core/surface syntax/typing figures

@@
 
+\newcommand{\figref}[1]{Figure \ref{fig:#1}}
+
 \inputott{surface.ott}
@@
 
+The syntax of the surface language is presented in \figref{surface_syntax}.
+
+The typing rules for the surface language are laid out in \figref{surface_typing}.
+For the type well-formedness and term typing rules
+we also assume each of them has a premise $\nonterm G \text{ ok}$
+stating that the context is well-formed.
+
 \begin{figure}[ht]
@@
   \footnotesize
+  \drules[TCTX]{$\nonterm G \text{ ok}$}{context well-formedness}{Empty,Bind}
   \drules[TWF]{$\nonterm G \vdash \nonterm{ts}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
   \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{Unit,Var,Abs,App,Case,Con,Sub}
+  \drules[ST]{$\nonterm G \vdash \nonterm{ts}_1 <: \nonterm{ts}_2$}{subtyping}{Base,Arr}
   \caption{Surface language typing}
   \label{fig:surface_typing}
-\end{figure}
-
-\begin{figure}[ht]
-  \footnotesize
-  \drules[ST]{$\nonterm G \vdash \nonterm{ts}_1 <: \nonterm{ts}_2$}{subtyping}{Base,Arr}
-  \caption{Surface language subtyping}
-  \label{fig:surface_subtyping}
 \end{figure}
@@
 does not break the usual metatheoretical properties of the system.
+
+The syntax of the core language is presented in \figref{core_syntax},
+and the typing rules are defined in \figref{core_typing}.
 

### 2020-07-19 tex/paper.tex
Add commands for |- translations

@@
 \newcommand{\tranty}{\mu_\tau}
-\newcommand{\tranderiv}{\mu_\vdash}
 \newcommand{\tranterm}{\mu_\varepsilon}
 
+\newcommand{\Tranctx}{\mu_{\vdash\Gamma}}
+\newcommand{\Tranty}{\mu_{\vdash\tau}}
+\newcommand{\Tranterm}{\mu_{\vdash\varepsilon}}
+
 We translate the surface language contexts, types and terms into the corresponding core language entities
-using the $\tranctx$, $\tranty$ and $\tranderiv$ functions respectively,
+using the $\tranctx$, $\tranty$ and $\tranterm$ functions respectively,
 which are defined mutually recursively.
 
-Before we dive into defining these functions, $\tranderiv$ deserves a few introductory words.
+Before we dive into defining these functions, $\Tranterm$ deserves a few introductory words.
 Although the aim is to translate terms,

### 2020-07-20 tex/paper.tex
Start writing the terms translation the right way

@@
 \paragraph{Terms.}
-Finally we define $\tranderiv$.
-As it is defined on typing derivation trees,
-we consider the term typing rule that is used last to prove $\nonterm G \vdash \nonterm{es} : \nonterm{ts}$.
+Finally we define $\Tranterm$.
+We consider the term typing rule that is at the root of the derivation tree for $\nonterm G \vdash \nonterm{es} : \nonterm{ts}$.
 
@@
   \item \ranchor{T-Unit}.
-    This rule has no premises, so we just translate it into \ranchor{CT-UnitTerm}:
-    \[
-      \inferrule{ }{\nonterm G \vdash \unit : \{ \gterm v : \Unit | \top \}} \longmapsto \inferrule{ }{\nonterm{GC} \vdash \unitc : \Unitc}
-    \]
+    \begin{align*}
+      & \inferrule{\nonterm G \text{ ok}}{\nonterm G \vdash \unit : \{ \gterm v : \Unit | \top \}}
+          \longmapsto
+        \inferrule{ }{\nonterm{GC} \vdash \unitc : \Unitc}, \\
+      \text{where}~&\nonterm{GC} = \tranctx(\nonterm G \text{ ok}).
+    \end{align*}
+    This rule has a single premise $\nonterm G \text{ ok}$ that the surface context is well-formed.
+    We use $\Tranctx$ to translate the surface derivation of that premise into the core one,
+    and extract the resulting context $\nonterm{GC}$.
+    We then use that context for \ranchor{CT-UnitTerm}.
 \end{itemize}
-
-\begin{equation}
-\begin{aligned}
-  & \tranterm(x) &&
-    \triangleq x                                                              \\
-  & \tranterm(\lambda x : \nonterm{ts}. \nonterm{es})  &&
-    \triangleq \lambda x : \tranty(\nonterm{ts}). \tranterm(\nonterm{es})     \\
-  & \tranterm(\nonterm{es}_1 \nonterm{es}_2)  &&
-    \triangleq \tranterm(\nonterm{es}_1) \tranterm(\nonterm{es}_2)            \\
-\end{aligned}
-\end{equation}
 

### 2020-07-20 tex/paper.tex
Add a nonterm for core types
Add a TWF rule for ADTs
Proof of well-typedness preservation becomes trivial at least

@@
 \newnonterm{ts}{\tau}
+\newnonterm{tc}{\hat\tau}
 \newnonterm s s
@@
   \drules[TCTX]{$\nonterm G \text{ ok}$}{context well-formedness}{Empty,Bind}
-  \drules[TWF]{$\nonterm G \vdash \nonterm{ts}$}{type well-formedness}{TrueRef,Base,Conj,Arr}
+  \drules[TWF]{$\nonterm G \vdash \nonterm{ts}$}{type well-formedness}{TrueRef,Base,Conj,Arr,Rec}
   \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{Unit,Var,Abs,App,Case,Con,Sub}
@@
   For any well-typed surface language term $\nonterm{es}$,
-  the result of the translation $\tranterm(\nonterm{es})$ is ``equivalently'' typeable in the core language:
-  \[
-    \nonterm G \vdash \nonterm{es} : \nonterm{ts} \implies \tranctx(\nonterm G) \vdash \tranterm(\nonterm{es}) : \tranty(\nonterm{ts})
-  \]
+  the result of the translation $\tranterm(\nonterm{es})$ is ``equivalently'' typeable in the core language.
 \end{theorem}
 \begin{proof}
-  We prove this by induction on the derivation of the typeability of $\nonterm{es}$.
-  In particular, we consider the last typing rule used:
+  This follows directly from the definition of $\tranterm$ via $\Tranterm$,
+  the latter being total and mapping onto typing derivations in the core language.
 \end{proof}

### 2020-07-20 tex/paper.tex
State the functions are total

@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
 
@@
 
+Having defined these functions, we formulate a simple but useful lemma:
+\begin{lemma}[Translation totality]
+  $\Tranctx$, $\Tranty$ and $\Tranterm$ are total.
+\end{lemma}
+\begin{proof}
+  Firstly, each function is defined for all possible inputs.
+  Then, each function recurses on structurally smaller inputs,
+  and, since derivation trees are finite,
+  this implies each function terminates for any input.
+\end{proof}
+
 \subsection{Correctness}

### 2020-07-20 tex/paper.tex
Add CT-Weaken derived rule

@@
 \begin{itemize}
+  \item Simplified variable typing rule which we call \ranchor{CT-VarW}:
+    \[
+      \inferrule{\nonterm{GC} \vdash \nonterm{ec} : s \\ \gterm x : \nonterm{ec} \in \nonterm{GC}}
+                {\nonterm{GC} \vdash \gterm x : \nonterm{ec}}.
+    \]
+    This rule unfolds into a single \ranchor{CT-Var} defining $x$,
+    followed by a sequence of \ranchor{CT-Weaken}
+    to ``move'' the $x$ into the right position in the final $\nonterm{GC}$.
   \item Non-dependent function type:

### 2020-07-20 tex/paper.tex
Better define \tranctx

@@
 We start with defining $\tranctx$,
-which merely translates the types of the bindings in the context from the surface language into the core language:
+which is the simplest translation of the three.
+It merely maps the types of the bindings in a (well-formed) context
+from the surface language into the core language.
+
 \begin{equation}
 \begin{aligned}
-  & \tranctx(\emptyset)                                       && \triangleq \emptyset                                               \\
-  & \tranctx(\nonterm G, \gterm x : \nonterm{ts})             && \triangleq \tranctx(\nonterm G), \gterm x : \tranty(\nonterm{ts})  \\
+  & \tranctx(\emptyset)
+      && \triangleq
+    \emptyset                                               \\
+  & \tranctx(\nonterm G, \gterm x : \nonterm{ts})
+      && \triangleq
+    \tranctx(\nonterm G), \gterm x : \tranty(\nonterm {G} \vdash \nonterm{ts})  \\
 \end{aligned}

### 2020-07-20 tex/paper.tex
We don't really \Tranctx that operates on `Γ ok`

@@
 
-\newcommand{\Tranctx}{\mu_{\vdash\Gamma}}
 \newcommand{\Tranty}{\mu_{\vdash\tau}}
@@
     This rule has a single premise $\nonterm G \text{ ok}$ that the surface context is well-formed.
-    We use $\Tranctx$ to translate the surface derivation of that premise into the core one,
+    We use $\tranctx$ to translate the surface derivation of that premise into the core one,
     and extract the resulting context $\nonterm{GC}$.
@@
 \begin{lemma}[Translation totality]
-  $\Tranctx$, $\Tranty$ and $\Tranterm$ are total.
+  $\tranctx$, $\Tranty$ and $\Tranterm$ are total.
 \end{lemma}

### 2020-07-20 tex/paper.tex
Properly define context mapping

@@
 \paragraph{Contexts.}
-We start with defining $\tranctx$,
+As a warm-up,
+we start with defining $\Tranctx$,
 which is the simplest translation of the three.
@@
 
-\begin{equation}
-\begin{aligned}
-  & \tranctx(\emptyset)
-      && \triangleq
-    \emptyset                                               \\
-  & \tranctx(\nonterm G, \gterm x : \nonterm{ts})
-      && \triangleq
-    \tranctx(\nonterm G), \gterm x : \tranty(\nonterm {G} \vdash \nonterm{ts})  \\
-\end{aligned}
-\end{equation}
+\begin{itemize}
+  \item \ranchor{TCTX-Empty}.
+    \begin{align*}
+      & \inferrule{~}
+                  {\emptyset \text{~ok}}
+          \longmapsto
+        \emptyset.
+    \end{align*}
+    The base case is trivial: we map the empty context to the empty context.
+  \item \ranchor{TCTX-Bind}.
+    \begin{align*}
+      & \inferrule{\nonterm G \text{~ok} \\ \nonterm G \vdash \nonterm{ts}}
+                  {\nonterm G , \gterm x : \nonterm{ts} \text{~ok}}
+          \longmapsto
+        \Tranctx(\nonterm G \text{~ok}), x : \tranty(\nonterm G \vdash \nonterm{ts}).
+    \end{align*}
+
+    We recurse on the prefix $\nonterm G$,
+    which we are allowed to do due to the $\nonterm G \text{~ok}$ premise,
+    and we use $\tranty$ to get the translated type of $\gterm x$,
+    which we are also allowed to do due to the corresponding premise.
+\end{itemize}
 

### 2020-07-20 tex/paper.tex
Build fix
Add \ctxok command to avoid code duplication

--- 2020-07-20 tex/paper.tex
@@
 \newcommand{\Int}{\textbf{Int}}
+\newcommand{\ctxok}{\text{~ok}}
 
@@
 For the type well-formedness and term typing rules
-we also assume each of them has a premise $\nonterm G \text{ ok}$
+we also assume each of them has a premise $\nonterm G \ctxok$
 stating that the context is well-formed.
@@
   \footnotesize
-  \drules[TCTX]{$\nonterm G \text{ ok}$}{context well-formedness}{Empty,Bind}
+  \drules[TCTX]{$\nonterm G \ctxok$}{context well-formedness}{Empty,Bind}
   \drules[TWF]{$\nonterm G \vdash \nonterm{ts}$}{type well-formedness}{TrueRef,Base,Conj,Arr,Rec}
@@
 
-\newcommand{\tranctx}{\mu_\Gamma}
 \newcommand{\tranty}{\mu_\tau}
@@
 
-\begin{equation}
-\begin{aligned}
-  & \tranctx(\emptyset)
-      && \triangleq
-    \emptyset                                               \\
-  & \tranctx(\nonterm G, \gterm x : \nonterm{ts})
-      && \triangleq
-    \tranctx(\nonterm G), \gterm x : \tranty(\nonterm {G} \vdash \nonterm{ts})  \\
-\end{aligned}
-\end{equation}
+\begin{itemize}
+  \item \ranchor{TCTX-Empty}.
+    \begin{align*}
+      & \inferrule{~}
+                  {\emptyset \ctxok}
+          \longmapsto
+        \emptyset.
+    \end{align*}
+    The base case is trivial: we map the empty context to the empty context.
+  \item \ranchor{TCTX-Bind}.
+    \begin{align*}
+      & \inferrule{\nonterm G \ctxok \\ \nonterm G \vdash \nonterm{ts}}
+                  {\nonterm G , \gterm x : \nonterm{ts} \ctxok}
+          \longmapsto
+        \Tranctx(\nonterm G \ctxok), x : \tranty(\nonterm G \vdash \nonterm{ts}).
+    \end{align*}
+
+    We recurse on the prefix $\nonterm G$,
+    which we are allowed to do due to the $\nonterm G \ctxok$ premise,
+    and we use $\tranty$ to get the translated type of $\gterm x$,
+    which we are also allowed to do due to the corresponding premise.
+\end{itemize}
 
@@
     \begin{align*}
-      & \inferrule{\nonterm G \text{ ok}}{\nonterm G \vdash \unit : \{ \gterm v : \Unit | \top \}}
+      & \inferrule{\nonterm G \ctxok}
+                  {\nonterm G \vdash \unit : \{ \gterm v : \Unit | \top \}}
           \longmapsto
-        \inferrule{ }{\nonterm{GC} \vdash \unitc : \Unitc}, \\
-      \text{where}~&\nonterm{GC} = \tranctx(\nonterm G \text{ ok}).
+        \inferrule{~}
+                  {\nonterm{GC} \vdash \unitc : \Unitc}, \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok).
     \end{align*}
-    This rule has a single premise $\nonterm G \text{ ok}$ that the surface context is well-formed.
+    This rule has a single premise $\nonterm G \ctxok$ that the surface context is well-formed.
     We use $\Tranctx$ to translate the surface derivation of that premise into the core one,

### 2020-07-20 tex/paper.tex
Throw up some introductory text on the translation functions

@@
 
-We translate the surface language contexts, types and terms into the corresponding core language entities
-using the $\tranctx$, $\tranty$ and $\tranterm$ functions respectively,
-which are defined mutually recursively.
+The surface language ultimately gets translated into the core language.
+What really matters for practical purposes is the translation of the terms since
+terms, not types or contexts, will eventually be compiled down to some executable code.
+On the other hand, it seems desirable to ensure that the terms
+that are well-typed in the surface language
+are accepted by the core type checker after the translation,
+since this provides some level of guarantee that the translation ``makes sense''.
 
-Before we dive into defining these functions, $\Tranterm$ deserves a few introductory words.
-Although the aim is to translate terms,
-what this function really does is translating the surface typing derivation for a given (well-typed) term
-into a typing derivation in the core language.
-The reason for the extra complexity stems from the \ranchor{T-Sub} typing rule.
-In particular, this rule invokes the subtyping check, which, in turn, invokes the oracle.
-The oracle then might produce a core language term proving that the subtyping relation in question holds,
-and the presence of this rule is in some sense a carrier of this proof term.
+In particular, note that the oracle plays a crucial role in typechecking of the surface language,
+so it is natural to expect that its output will also be used during any such translation,
+for instance, to produce core language proofs that VCs hold.
+The only trace of the oracle's work is in the typing derivation
+(namely, in the subtyping relation check and the \ranchor{T-Sub} typing rule),
+so we choose to do the translation not on \emph{terms},
+but rather on \emph{well-typedness derivations} for the terms.
+This way the oracle has a chance to enrich the original surface language term
+with proofs and witnesses of whatever it decided.
+
+Thus the ultimate goal of this section is to define a function $\Tranterm$
+taking a typing derivation in the surface language
+and producing a typing derivation in the core language.
+As part of its duty
+this function also needs to translate the types and contexts
+encountered in the source derivation,
+so we also define two helper functions.
+$\Tranty$ takes a surface derivation of a \emph{type} well-formedness
+and produces the corresponding core derivation.
+$\Tranctx$ takes a surface derivation of a \emph{context} well-formedness
+and produces a core language context.
+
+We also define two helper functions, $\tranty$ and $\tranterm$,
+invoking $\Tranty$ and $\Tranterm$ and extracting just the type and term component respectively.
+
+All in all, our goal is to define $\Tranctx$, $\Tranty$ and $\Tranterm$
+with ``metatypes''
+\begin{align*}
+  \Tranctx  & : \nonterm G \ctxok                             \longmapsto \nonterm{GC},                                     \\
+  \Tranty   & : \nonterm G \vdash \nonterm{ts}                \longmapsto \nonterm{GC} \vdash \nonterm{tc} : \star,         \\
+  \Tranterm & : \nonterm G \vdash \nonterm{es} : \nonterm{ts} \longmapsto \nonterm{GC} \vdash \nonterm{ec} : \nonterm{tc},
+\end{align*}
+Our helper functions $\tranty$ and $\tranterm$ are then defined to return the $\nonterm{tc}$ and $\nonterm{ec}$
+from the right-hand sides above respectively.
+
+The apparent asymmetry of $\Tranctx$, which does not produce a derivation as others do,
+is due to our core language not having a separate notion of a context well-formedness.
 

### 2020-07-20 tex/paper.tex
Better context translation description

@@
 we start with defining $\Tranctx$,
-which is the simplest translation of the three.
+which is the simplest of the three.
 It merely maps the types of the bindings in a (well-formed) context
@@
 
+Since $\Tranctx$ is defined on the derivations of $\nonterm G \ctxok$,
+we consider the last rule used in a derivation:
 \begin{itemize}
@@
     We recurse on the prefix $\nonterm G$,
-    which we are allowed to do due to the $\nonterm G \ctxok$ premise,
+    which is admissible due to the $\nonterm G \ctxok$ premise,
     and we use $\tranty$ to get the translated type of $\gterm x$,
-    which we are also allowed to do due to the corresponding premise.
+    which is also admissible due to the other premise.
 \end{itemize}

### 2020-07-20 tex/paper.tex
Mention the right vars in translation totality lemma
Add a T-Var clause for \Tranterm

@@
     We then use that context for \ranchor{CT-UnitTerm}.
+
+  \item \ranchor{T-Var}.
+    \begin{align*}
+      & \inferrule{\nonterm G \ctxok \\
+                   \gterm x : \nonterm{ts} \in \nonterm G}
+                  {\nonterm G \vdash \gterm x : \nonterm{ts}}
+          \longmapsto
+        \inferrule{x : \nonterm{tc} \in \nonterm{GC}}
+                  {\nonterm{GC} \vdash \gterm x : \nonterm{tc}},    \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok),       \\
+                   &\nonterm{tc} = \tranty(\Gamma \vdash \nonterm{ts}).
+    \end{align*}
+
+    To show that the right-hand side is well-defined,
+    we need to show that $x : \nonterm{tc} \in \nonterm{GC}$ holds.
+
+    By definition of $\Tranctx$,
+    if $\nonterm G$ contains a binding for the name $x$,
+    then $\nonterm{GC}$ also contains a binding for that name,
+    and its type is precisely $\tranty(\Gamma \vdash \nonterm{ts})$.
+    Thus, we are allowed to use the (derived) core typing rule \ranchor{CT-VarW}.
 \end{itemize}
@@
 \begin{lemma}[Translation totality]
-  $\tranctx$, $\Tranty$ and $\Tranterm$ are total.
+  $\Tranctx$, $\Tranty$ and $\Tranterm$ are total.
 \end{lemma}

### 2020-07-20 tex/paper.tex
Add some facts about (non)dependent pairs

@@
 
-We also establish some derived forms in the core language to simplify the subsequent exposition:
-
+We also establish some derived forms in the core language to simplify the subsequent exposition,
+and state some straightforward facts that will be useful later:
 \begin{itemize}
@@
     \]
+    It's easy to see from the formation rule that
+    \begin{lemma}
+      If $\nonterm{tc}$ and $P x$ are well-formed in $\nonterm{GC}$,
+      then $(x : \nonterm{tc} \dast P x)$ is well-formed in $\nonterm{GC}$ and has type $\star$.
+    \end{lemma}
   \item Non-dependent pair type:
@@
     \]
+    As a direct consequence of the previous lemma we obtain
+    \begin{lemma}
+      If $\nonterm{tc}_1, \nonterm{tc}_2$ are well-formed in $\nonterm{GC}$,
+      then $(\nonterm{tc}_1, \nonterm{tc}_2)$ is well-formed in $\nonterm{GC}$ and has type $\star$.
+    \end{lemma}
   \item Equality of expressions $\nonterm{ec}_1, \nonterm{ec}_2$ of type $S$:

### 2020-07-21 tex/paper.tex
More consistent \tau usage in core language
Lemma hyperrefing
Make the non-dependent pair labeled

@@
 \usepackage{amsmath,amsfonts,amssymb,amsthm}
+\usepackage{hyperref,cleveref}
 \usepackage{mathtools}
@@
 \newtheorem{lemma}{Lemma}
+\crefname{lemma}{Lemma}{Lemmas}
 
@@
 
-Although the syntax doesn't mention $\tau$, we will use it as a metavariable
+Although the syntax doesn't mention $\nonterm{tc}$, we will use it as a metavariable
 where a type (an inhabitant of $\star$) is expected.
@@
     \[
-      \tau_1 \rightarrow \tau_2 \triangleq \Pi \_ : \tau_1. \tau_2.
+      \nonterm{tc}_1 \rightarrow \nonterm{tc}_2 \triangleq \Pi \_ : \nonterm{tc}_1. \nonterm{tc}_2.
     \]
@@
     \[
-      (x : \tau \dast P x) \triangleq \Pi \alpha : \star. (\Pi x : \tau. P x \rightarrow \alpha) \rightarrow \alpha,
+      (x : \nonterm{tc} \dast P x) \triangleq \Pi \alpha : \star. (\Pi x : \nonterm{tc}. P x \rightarrow \alpha) \rightarrow \alpha,
     \]
@@
     \[
-      (x \dast p) \triangleq \lambda \alpha : \star. \lambda f : (\Pi x' : \tau. P x' \rightarrow \alpha). f x p
+      (x \dast p) \triangleq \lambda \alpha : \star. \lambda f : (\Pi x' : \nonterm{tc}. P x' \rightarrow \alpha). f x p
     \]
@@
     \[
-      \text{fst} \triangleq \lambda \sigma : (x : \tau \dast P x). \sigma \tau (\lambda x' : \tau. \lambda \_ : P x'. x'),
+      \text{fst} \triangleq \lambda \sigma : (x : \nonterm{tc} \dast P x). \sigma \nonterm{tc} (\lambda x' : \nonterm{tc}. \lambda \_ : P x'. x'),
     \]
     \[
-      \text{snd} \triangleq \lambda \sigma : (x : \tau \dast P x). \sigma \tau (\lambda x' : \tau. \lambda p : P x'. p).
+      \text{snd} \triangleq \lambda \sigma : (x : \nonterm{tc} \dast P x). \sigma \nonterm{tc} (\lambda x' : \nonterm{tc}. \lambda p : P x'. p).
     \]
@@
     \[
-      (\tau_1, \tau_2) \triangleq \{ x : \tau_1\ **\ (\lambda \_ : \tau_1. \tau_2) x \}.
+      (\nonterm{tc}_1, \nonterm{tc}_2) \triangleq \{ x : \nonterm{tc}_1\ **\ (\lambda \_ : \nonterm{tc}_1. \nonterm{tc}_2) x \}.
     \]
     As a direct consequence of the previous lemma we obtain
-    \begin{lemma}
+    \begin{lemma}\label{lma:non_dep_pair_typing}
       If $\nonterm{tc}_1, \nonterm{tc}_2$ are well-formed in $\nonterm{GC}$,

### 2020-07-21 tex/paper.tex
Add a clause for TWF-TrueRef

@@
 \paragraph{Types.}
-$\tranty$ is defined as follows,
+Next, we define $\Tranty$.
+We now consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$.
+
+\begin{itemize}
+  \item \ranchor{TWF-TrueRef}.
+    \begin{align*}
+      & \inferrule{\nonterm G \ctxok}
+                  {\nonterm G \vdash \{ v : B | \top \}}
+          \longmapsto
+        \inferrule{\cdots}
+                  {\nonterm{GC} \vdash (\hat B, \unitc \equiv \unitc) : *}, \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok).
+    \end{align*}
+
+    Here we somewhat frivolously use $\hat B$ to denote
+    the core language analogue of the surface language base type $B$.
+    For example, if this rule is instantiated for $B$ equal to $\Unit$,
+    then $\hat B$ would be $\Unitc$.
+
+    We also do not mention the premises of the translated rule.
+    By \cref{lma:non_dep_pair_typing}
+    $(\hat B, \unitc \equiv \unitc)$ has type $\star$ in $\nonterm{GC}$
+    if $\hat B$ and $\unitc \equiv \unitc$ are well-typed in $\nonterm{GC}$,
+    which, in turn, is easily checked.
+    Thus, there exists a derivation of $\nonterm{GC} \vdash (\hat B, \unitc \equiv \unitc) : *$,
+    and it is effectively constructible~--- we just omit it for brevity.
+\end{itemize}
+
+$\Tranty$ is defined as follows,
 with the slight abuse of notation to allow ``recursive'' calls

### 2020-07-21 tex/paper.tex
Add a clause for TWF-Arr

@@
     and it is effectively constructible~--- we just omit it for brevity.
+  \item \ranchor{TWF-Arr}.
+    \begin{align*}
+      & \inferrule{\nonterm G \ctxok \\\\
+                   \nonterm G \vdash \nonterm{ts}_1 \\
+                   \nonterm G , \gterm x : \nonterm{ts}_1 \vdash \nonterm{ts}_2}
+                  {\nonterm G \vdash (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2}
+          \longmapsto
+        \inferrule{\Tranty(\nonterm G \vdash \nonterm{ts}_1) \\
+                   \Tranty(\nonterm G , \gterm x : \nonterm{ts}_1 \vdash \nonterm{ts}_2)}
+                  {\nonterm{GC} \vdash \Pi x : \nonterm{tc}_1. \nonterm{tc}_2 : \star}, \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok),                                         \\
+                   &\nonterm{tc}_1 = \tranty(\nonterm G \vdash \nonterm{ts}_1),                         \\
+                   &\nonterm{tc}_2 = \tranty(\nonterm G , \gterm x : \nonterm{ts}_1 \vdash \nonterm{ts}_2).
+    \end{align*}
+
+    We resort to \ranchor{CT-Form} on the right.
+    This is well-formed,
+    since $\Tranty(\nonterm G \vdash \nonterm{ts}_1)$ produces a derivation of $\nonterm{GC} \vdash \nonterm{tc}_1 : \star$,
+    and similarly we get a derivation of $\nonterm{GC}, \gterm x : \nonterm{tc}_1 \vdash \nonterm{tc}_2 : \star$ from
+    the other invocation of $\Tranty$.
 \end{itemize}

### 2020-07-22 tex/paper.tex
Add a separate \fresh shortcut
Minor text style fix
Use \textsc instead of \ranchor

@@
 \newcommand{\ctxok}{\text{~ok}}
+\newcommand{\fresh}{\text{~fresh name}}
 
@@
 \begin{itemize}
-  \item Simplified variable typing rule which we call \ranchor{CT-VarW}:
+  \item Simplified variable typing rule which we call \textsc{CT-VarW}:
     \[
@@
     \]
-    This rule unfolds into a single \ranchor{CT-Var} defining $x$,
-    followed by a sequence of \ranchor{CT-Weaken}
+    This rule unfolds into a single \textsc{CT-Var} defining $x$,
+    followed by a sequence of \textsc{CT-Weaken}
     to ``move'' the $x$ into the right position in the final $\nonterm{GC}$.
@@
 The only trace of the oracle's work is in the typing derivation
-(namely, in the subtyping relation check and the \ranchor{T-Sub} typing rule),
+(namely, in the subtyping relation check and the \textsc{T-Sub} typing rule),
 so we choose to do the translation not on \emph{terms},
@@
 \begin{itemize}
-  \item \ranchor{TCTX-Empty}.
+  \item \textsc{TCTX-Empty}.
     \begin{align*}
@@
     The base case is trivial: we map the empty context to the empty context.
-  \item \ranchor{TCTX-Bind}.
+  \item \textsc{TCTX-Bind}.
     \begin{align*}
@@
 \begin{itemize}
-  \item \ranchor{TWF-TrueRef}.
+  \item \textsc{TWF-TrueRef}.
     \begin{align*}
@@
     and it is effectively constructible~--- we just omit it for brevity.
-  \item \ranchor{TWF-Arr}.
+  \item \textsc{TWF-Arr}.
     \begin{align*}
@@
 
-    We resort to \ranchor{CT-Form} on the right.
+    We resort to \textsc{CT-Form} on the right.
     This is well-formed,
@@
 Finally we define $\Tranterm$.
-We consider the term typing rule that is at the root of the derivation tree for $\nonterm G \vdash \nonterm{es} : \nonterm{ts}$.
+We consider the term typing rule at the root of the derivation tree for $\nonterm G \vdash \nonterm{es} : \nonterm{ts}$.
 
 \begin{itemize}
-  \item \ranchor{T-Unit}.
+  \item \textsc{T-Unit}.
     \begin{align*}
@@
     and extract the resulting context $\nonterm{GC}$.
-    We then use that context for \ranchor{CT-UnitTerm}.
+    We then use that context for \textsc{CT-UnitTerm}.
 
-  \item \ranchor{T-Var}.
+  \item \textsc{T-Var}.
     \begin{align*}
@@
     and its type is precisely $\tranty(\Gamma \vdash \nonterm{ts})$.
-    Thus, we are allowed to use the (derived) core typing rule \ranchor{CT-VarW}.
+    Thus, we are allowed to use the (derived) core typing rule \textsc{CT-VarW}.
 \end{itemize}

### 2020-07-22 tex/paper.tex
Remove the old-style tranty definition

@@
 
-$\Tranty$ is defined as follows,
-with the slight abuse of notation to allow ``recursive'' calls
-on the refinement part of the surface language syntax:
-\begin{equation}
-\begin{aligned}
-  & \tranty( (x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2)  && \triangleq \Pi x : \tranty(\nonterm{ts}_1). \tranty(\nonterm{ts}_2)     \\
-  & \tranty( \{ v : \nonterm{ts} | \nonterm{r} \})             && \triangleq \{ v : \tranty(\nonterm{ts})\ **\ \tranty(\nonterm r) \}     \\
-  & \tranty( \nonterm r_1 \land \nonterm r_2 )                 && \triangleq ( \tranty(\nonterm r_1) , \tranty(\nonterm r_2))             \\
-  & \tranty( \nonterm{es}_1 = \nonterm{es}_2 )                 && \triangleq \tranterm(\nonterm{es}_1) \equiv \tranterm(\nonterm{es}_2)   \\
-  & \tranty(B)                                                 && \triangleq B                                                            \\
-%TODO this last one is not well-defined
-\end{aligned}
-\end{equation}
 Or, in words:

### 2020-07-22 tex/paper.tex
Move the "in words" part before the formal definition of tranty
Remove the attempt at extending Boehm-Berarducci to dependent pattern-matching

@@
     The reader might recognize this as the Leibniz equality \cite{FindSomethingForThis}.
-  \item ADTs, expressed using an encoding similar to Boehm-Berarducci's \cite{Bohm85}.
-    We borrow the syntax for the ADTs from the surface language and put a hat on it,
-    so let's take some ADT $\widehat{\langle \overline{l_i : \nonterm{ts}_i}^i \rangle}$.
-    The usual Boehm-Berarducci encoding associates the type
-    $\Pi \alpha : \star. \overline{(\nonterm{ts}_i \rightarrow \alpha)}^i \rightarrow \alpha$ with it.
-
-    We go a step further to account for the subset of the dependent pattern matching
-    that we are interested in.
-
-    First, each label $l_i$ gives rise to a term $\hat{l_i}$ of type $\nonterm{ts}_i \rightarrow \xi$.
-    %TODO operational semantics?
-    We then define the derived form
-    \[
-      \widehat{\langle \overline{l_i : \nonterm{ts}_i}^i \rangle}
-        \triangleq
-          \Pi s : \xi.
-          \Pi \alpha : \star.
-          \overline{(\Pi x_i : \nonterm{ts}_i. s \equiv \hat{l_i} x_i \rightarrow \alpha)}^i \rightarrow \alpha.
-    \]
-    The outer extra argument named $s$ denotes the scrutinee of the case expression to be defined next.
-    The inner extra argument with the type $s \equiv \hat{l_i} x_i$
-    expresses the subset of dependent pattern matching
-    that we rely on for the translation from the surface language:
-    in particular, it provides the ``case branch'' with a proof
-    that the scrutinee $s$ is indeed equal to the ``case pattern'' $\hat{l_i} x_i$.
 \end{itemize}
@@
 \paragraph{Types.}
-Next, we define $\Tranty$.
-We now consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$.
+Next, we define $\Tranty$. In short:
+\begin{itemize}
+  \item (dependent) arrow types are translated to the corresponding $\Pi$-types,
+  \item refined types are translated to the corresponding $\Sigma$-types,
+  \item atomic refinements, being propositions about surface language terms equality,
+    are translated to propositions stating the equality of translated core language terms,
+  \item conjunctions of several (atomic) refinements are represented as tuples of the corresponding atomic translations.
+\end{itemize}
+
+More formally, we consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$.
 
@@
     the other invocation of $\Tranty$.
-\end{itemize}
-
-Or, in words:
-\begin{itemize}
-  \item (dependent) arrow types are translated to the corresponding $\Pi$-types,
-  \item refined types are translated to the corresponding $\Sigma$-types,
-  \item atomic refinements, being propositions about surface language terms equality,
-    are translated to propositions stating the equality of translated core language terms,
-  \item conjunctions of several (atomic) refinements are represented as tuples of the corresponding atomic translations.
 \end{itemize}

### 2020-07-22 tex/paper.tex
Add a syntax for remarks

@@
 \newtheorem{lemma}{Lemma}
+\newtheorem{remark}{Remark}
 \crefname{lemma}{Lemma}{Lemmas}
+\crefname{remark}{Remark}{Remarks}
 
@@
 
+\begin{remark}\label{remark:surface_base_types}
   The surface language might also have other base types (for example, $\Int$),
@@
   but we omit them for the sake of brevity.
+\end{remark}
 

### 2020-07-22 tex/paper.tex
Lift out the remark for \hat B

@@
 
+In addition to that, we will somewhat frivolously use $\hat B$
+to denote the core language analogue of a surface language base type $B$,
+which exists for $\Unit$ ($\hat B$ then being $\Unitc$)
+and which we assume to exist for any other surface language base type
+mentioned in \cref{remark:surface_base_types}.
+
 \newcommand{\dast}{\ **\ }
@@
 
-    Here we somewhat frivolously use $\hat B$ to denote
-    the core language analogue of the surface language base type $B$.
-    For example, if this rule is instantiated for $B$ equal to $\Unit$,
-    then $\hat B$ would be $\Unitc$.
-
-    We also do not mention the premises of the translated rule.
+    We do not mention the premises of the translated rule.
     By \cref{lma:non_dep_pair_typing}

### 2020-07-22 tex/paper.tex
Fix the formulation of the dependent pair lemma

@@
     \]
-    It's easy to see from the formation rule that
+    It is easy to see from the formation rule that
     \begin{lemma}
-      If $\nonterm{tc}$ and $P x$ are well-formed in $\nonterm{GC}$,
-      then $(x : \nonterm{tc} \dast P x)$ is well-formed in $\nonterm{GC}$ and has type $\star$.
+      If $\nonterm{tc}$ is typeable in $\nonterm{GC}$
+      and $P x$ is typeable in $\nonterm{GC} , x : \nonterm{tc}$,
+      then $(x : \nonterm{tc} \dast P x)$ is also typeable in $\nonterm{GC}$ and,
+      in particular, it has type $\star$.
     \end{lemma}

### 2020-07-22 tex/paper.tex
Make the required lemmas named

@@
     It is easy to see from the formation rule that
-    \begin{lemma}
+    \begin{lemma}\label{lma:dep_pair_typing}
       If $\nonterm{tc}$ is typeable in $\nonterm{GC}$
@@
     \begin{lemma}\label{lma:non_dep_pair_typing}
-      If $\nonterm{tc}_1, \nonterm{tc}_2$ are well-formed in $\nonterm{GC}$,
-      then $(\nonterm{tc}_1, \nonterm{tc}_2)$ is well-formed in $\nonterm{GC}$ and has type $\star$.
+      If $\nonterm{tc}_1, \nonterm{tc}_2$ are typeable in $\nonterm{GC}$,
+      then $(\nonterm{tc}_1, \nonterm{tc}_2)$ is typeable in $\nonterm{GC}$ and has type $\star$.
     \end{lemma}
-  \item Equality of expressions $\nonterm{ec}_1, \nonterm{ec}_2$ of type $S$:
+  \item Equality of expressions $\nonterm{ec}_1, \nonterm{ec}_2$ of type $\nonterm{tc}$:
     \[
-      \nonterm{ec}_1 \equiv \nonterm{ec}_2 \triangleq \Pi P : S \rightarrow \star. (P x \rightarrow P y, P y \rightarrow P x).
+      \nonterm{ec}_1 \equiv \nonterm{ec}_2 \triangleq \Pi P : \nonterm{tc} \rightarrow \star. (P x \rightarrow P y, P y \rightarrow P x).
     \]
     The reader might recognize this as the Leibniz equality \cite{FindSomethingForThis}.
+
+    Again, it can be seen that
+    \begin{lemma}\label{lma:equality_typing}
+      If $\nonterm{ec}_1, \nonterm{ec}_2$ are well-typed in a context $\nonterm{GC}$,
+      then $\nonterm{GC} \vdash \nonterm{ec}_1 \equiv \nonterm{ec}_2 : \star$.
+    \end{lemma}
 \end{itemize}

### 2020-07-23 tex/paper.tex
Add TWF-Base case

@@
     and it is effectively constructible~--- we just omit it for brevity.
+  \item \textsc{TWF-Base}.
+    \begin{align*}
+      & \inferrule{\nonterm G \ctxok \\\\
+                   \nonterm G ' \vdash \nonterm{es}_1 : \{ v_2 : B' | \nonterm{r}_1 \} \\\\
+                   \nonterm G ' \vdash \nonterm{es}_2 : \{ v_2 : B' | \nonterm{r}_2 \} \\\\
+                   \gterm x \fresh}
+                  {\nonterm G \vdash \{ v : B | \nonterm{es}_1 = \nonterm{es}_2 \}}
+          \longmapsto
+        \inferrule{\Tranterm(\nonterm G ' \vdash \nonterm{es}_1 : \{ v_2 : B' | \nonterm{r}_1 \}) \\\\
+                   \Tranterm(\nonterm G ' \vdash \nonterm{es}_2 : \{ v_2 : B' | \nonterm{r}_2 \}) \\\\
+                   \gterm x \fresh \\\\
+                   \cdots}
+                  {\nonterm{GC} \vdash (v : \hat B \dast \lambda v : \hat B. \nonterm{ec}_1 \equiv \nonterm{ec}_2) : \star}, \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok), \\
+                   &\nonterm G ' = \nonterm G , v : \{ v_1 : B | \top \}, \\
+                   &\nonterm{ec}_i =
+                     \tranterm(\nonterm G , x : \{ v_1 : B | \top \} \vdash \nonterm{es}_i : \{ v_2 : B' | \nonterm{r}_i \}).
+    \end{align*}
+
+    To see that this is well-defined, first note that
+    that $\hat B$ has type $\star$ by assumption on $\hat B$.
+    Then, $\nonterm{ec}_1$ and $\nonterm{ec}_2$ are well-typed in the translated context $\nonterm G'$ according to the premises,
+    and this context coincides with the one used to type check $\lambda v : \hat B. \nonterm{ec}_1 \equiv \nonterm{ec}_2$.
+    Hence both components of the dependent pair are well-typed, so by \cref{lma:dep_pair_typing} the dependent pair itself has type $\star$.
+
+    Just as in the previous case, the extra derivations and premises corresponding to the considerations above
+    are omitted and denoted by $\cdots$.
   \item \textsc{TWF-Arr}.

### 2020-07-23 tex/paper.tex
Add a %TODO about the context not actually coinciding
Reuse Γ in definition of \hat e_i
TWF-Conj translation defined

@@
                    &\nonterm{ec}_i =
-                     \tranterm(\nonterm G , x : \{ v_1 : B | \top \} \vdash \nonterm{es}_i : \{ v_2 : B' | \nonterm{r}_i \}).
+                     \tranterm(\nonterm G ' \vdash \nonterm{es}_i : \{ v_2 : B' | \nonterm{r}_i \}).
     \end{align*}
@@
 
+    %TODO the { v : B | T } in the Γ' desugars to a dependent pair, yet we use λv : \hat B — note the naked base type.
+    %Figure out how to best fix this.
+
     Just as in the previous case, the extra derivations and premises corresponding to the considerations above
     are omitted and denoted by $\cdots$.
+  \item \textsc{TWF-Conj}.
+    \begin{align*}
+      & \inferrule{\nonterm G \ctxok \\\\
+                   \nonterm G \vdash \{ v : B | \nonterm{r}_1 \} \\
+                   \nonterm G \vdash \{ v : B | \nonterm{r}_2 \}}
+                  {\nonterm G \vdash \{ v : B | \nonterm{r}_1 \land \nonterm{r}_2 \}}
+          \longmapsto
+        \inferrule{\cdots \\\\
+                   \Tranty(\nonterm G \vdash \{ v : B | \nonterm{r}_1 \}) \\
+                   \Tranty(\nonterm G \vdash \{ v : B | \nonterm{r}_2 \})}
+                  {\nonterm{GC} \vdash (v : \hat B \dast \lambda v : \hat B. (P_1 v, P_2 v)) : \star}, \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok),                                         \\
+                   &P_1 \text{~s.t.~} (v : \hat B \dast \lambda v : \hat B. P_1 v) = \tranty(\nonterm G \vdash \{ v : B | \nonterm{r}_1 \}),  \\
+                   &P_2 \text{~s.t.~} (v : \hat B \dast \lambda v : \hat B. P_2 v) = \tranty(\nonterm G \vdash \{ v : B | \nonterm{r}_2 \}).
+    \end{align*}
+
+    The proof of the right-hand side being well-defined goes similarly to the cases considered previously
+    with an extra step.
+    Namely, note that the last two ``patterns'' on the left-hand side are irrefutable:
+    it can be seen by direct inspection and inductive reasoning that $\tranty$ always yields a pair
+    for a derivation of the form $\nonterm G \vdash \{ v : B | \nonterm{r} \}$.
+    Thus we can take the predicate components $P_1, P_2$ out from the corresponding dependent pairs
+    and repack them obtaining $\lambda v : \hat B. (P_1 v, P_2 v)$.
   \item \textsc{TWF-Arr}.

### 2020-07-23 tex/paper.tex
Add a note that we assume \hat B : \star
Minor style stuff

@@
 mentioned in \cref{remark:surface_base_types}.
+We also assume that $\hat B$ has type $\star$ without any extra premises.
 
@@
 
-More formally, we consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$.
-
+More formally, we consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$:
 \begin{itemize}
@@
 
-    The proof of the right-hand side being well-defined goes similarly to the cases considered previously
-    with an extra step.
+    The proof of the right-hand side being well-defined goes similarly to the cases considered previously,
+    but with an extra step.
     Namely, note that the last two ``patterns'' on the left-hand side are irrefutable:

### 2020-07-24 tex/paper.tex
Some yak shaving for TODOs
Formulate some lemmas about the core language

@@
 \usepackage{cite,url,babelbib}
+\usepackage{todonotes}
 \usepackage[english]{babel}
@@
 (otherwise it might be left as a separate lemma to be filled out by the user in the core language, or rejected altogether).
+
+\subsubsection{Properties}
+
+\begin{lemma}
+  If $\nonterm G \vdash \nonterm{es} : \nonterm{ts}$, then $\nonterm G \ctxok$.
+\end{lemma}
+\begin{proof}
+  \todo[inline]{Do the proof and remove the assumption $\nonterm G \ctxok$ from all the rules unless needed.}
+\end{proof}
+
+\begin{lemma}
+  If $\nonterm G \vdash \nonterm{es} : \nonterm{ts}$, then $\nonterm G \vdash \nonterm{ts}$.
+\end{lemma}
+\begin{proof}
+  \todo[inline]{Do the proof.}
+\end{proof}
 
@@
 
-    %TODO the { v : B | T } in the Γ' desugars to a dependent pair, yet we use λv : \hat B — note the naked base type.
-    %Figure out how to best fix this.
+    \todo[inline]{
+      The ${ v : B | \top }$ in the $\nonterm G '$ desugars to a dependent pair, yet we use $\lambda v : \hat B$ — note the naked base type.
+      Figure out how to best fix this.
+    }
 
@@
     the other invocation of $\Tranty$.
+  \item \textsc{TWF-Rec}.
+    \todo[inline]{Don't forget to define this.}
 \end{itemize}

### 2020-07-24 tex/paper.tex
Remove unused nonterm
Separate core λC typing rules and their extensions

@@
 \newnonterm{GC}{\hat\Gamma}
-\newnonterm{binop}{\odot}
 
@@
 The syntax of the core language is presented in \figref{core_syntax},
-and the typing rules are defined in \figref{core_typing}.
+the core $\lambda C$ typing rules are defined in \figref{core_typing}
+and the rules for our extensions are defined in \figref{core_typing_exts}.
 
@@
   \footnotesize
-  \drules[CT]{$\nonterm G \vdash \nonterm{ec}_1 : \nonterm{ec}_2$}{core typing}{Sort,UnitType,UnitTerm,Var,Weaken,Form,App,Abs,Conv}
-  \caption{Core language typing rules}
+  \drules[CT]{$\nonterm G \vdash \nonterm{ec}_1 : \nonterm{ec}_2$}{core typing}{Sort,Var,Weaken,Form,App,Abs,Conv}
+  \caption{$\lambda C$ typing rules}
   \label{fig:core_typing}
+\end{figure}
+\begin{figure}[ht]
+  \footnotesize
+  \drules[CT]{$\nonterm G \vdash \nonterm{ec}_1 : \nonterm{ec}_2$}{core typing, extensions}{UnitType,UnitTerm}
+  \caption{Our extensions typing rules}
+  \label{fig:core_typing_exts}
 \end{figure}

### 2020-07-24 tex/paper.tex
Add ADTs to the core language
Sigh.
Add a note on our extensions and the usage of equality

@@
 \newgterm n n
+\newgterm p \pi
 
@@
 We take $\lambda C$ \cite{TTFPLambdaC}
-with a unit type as our core dependently typed language.
-We conjecture that the introduction of units
-does not break the usual metatheoretical properties of the system.
+extended with a unit type
+and a restricted form of ADTs as our core dependently typed language.
+We conjecture that these extensions
+do not break the usual metatheoretical properties of the system,
+and we explicitly prove progress and preservation theorems as a sanity check.
 
@@
 and the rules for our extensions are defined in \figref{core_typing_exts}.
+Note that the \textsc{CT-ADTCase} rule uses the equality $\equiv$,
+which is a derived form defined later.
 
@@
   \footnotesize
-  \drules[CT]{$\nonterm G \vdash \nonterm{ec}_1 : \nonterm{ec}_2$}{core typing, extensions}{UnitType,UnitTerm}
-  \caption{Our extensions typing rules}
+  \drules[CT]{$\nonterm G \vdash \nonterm{ec}_1 : \nonterm{ec}_2$}{core typing, extensions}{UnitType,UnitTerm,ADTForm,ADTCon,ADTCase}
+  \caption{Typing rules for our extensions to $\lambda C$}
   \label{fig:core_typing_exts}

### 2020-07-25 tex/paper.tex
Rename TWF-Rec → TWF-ADT for consistency
More consistent \star usage

@@
   \drules[TCTX]{$\nonterm G \ctxok$}{context well-formedness}{Empty,Bind}
-  \drules[TWF]{$\nonterm G \vdash \nonterm{ts}$}{type well-formedness}{TrueRef,Base,Conj,Arr,Rec}
+  \drules[TWF]{$\nonterm G \vdash \nonterm{ts}$}{type well-formedness}{TrueRef,Base,Conj,Arr,ADT}
   \drules[T]{$\nonterm G \vdash \nonterm{es} : \nonterm{ts}$}{term typing}{Unit,Var,Abs,App,Case,Con,Sub}
@@
         \inferrule{\cdots}
-                  {\nonterm{GC} \vdash (\hat B, \unitc \equiv \unitc) : *}, \\
+                  {\nonterm{GC} \vdash (\hat B, \unitc \equiv \unitc) : \star}, \\
       \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok).
@@
     which, in turn, is easily checked.
-    Thus, there exists a derivation of $\nonterm{GC} \vdash (\hat B, \unitc \equiv \unitc) : *$,
+    Thus, there exists a derivation of $\nonterm{GC} \vdash (\hat B, \unitc \equiv \unitc) : \star$,
     and it is effectively constructible~--- we just omit it for brevity.
@@
     the other invocation of $\Tranty$.
-  \item \textsc{TWF-Rec}.
+  \item \textsc{TWF-ADT}.
     \todo[inline]{Don't forget to define this.}

### 2020-07-25 tex/paper.tex
Define the translation for TWF-ADT
More consistent \nonterm G usage

@@
   \item \textsc{TWF-ADT}.
-    \todo[inline]{Don't forget to define this.}
+    \begin{align*}
+      & \inferrule{\nonterm G \ctxok \\\\
+                   \overline{\nonterm G \vdash \nonterm{ts}_i}^i
+                  }
+                  {\nonterm G \vdash \langle \overline{l_i : \nonterm{ts}_i}^i \rangle}
+          \longmapsto
+        \inferrule{\overline{\Tranty(\nonterm G \vdash \nonterm{ts}_i)}^i}
+                  {\nonterm{GC} \vdash \langle \overline{l_i : \nonterm{tc}_i}^i \rangle : \star}, \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok),                                         \\
+                   &\nonterm{tc}_i = \tranty(\nonterm G \vdash \nonterm{ts}_i).
+    \end{align*}
+
+    It can be seen that the right-hand side is well-defined
+    by an argument similar to ones for the previous clauses.
 \end{itemize}
@@
       \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok),       \\
-                   &\nonterm{tc} = \tranty(\Gamma \vdash \nonterm{ts}).
+                   &\nonterm{tc} = \tranty(\nonterm G \vdash \nonterm{ts}).
     \end{align*}
@@
     then $\nonterm{GC}$ also contains a binding for that name,
-    and its type is precisely $\tranty(\Gamma \vdash \nonterm{ts})$.
+    and its type is precisely $\tranty(\nonterm G \vdash \nonterm{ts})$.
     Thus, we are allowed to use the (derived) core typing rule \textsc{CT-VarW}.

### 2020-07-25 tex/paper.tex
Minor wording fix

@@
 Our helper functions $\tranty$ and $\tranterm$ are then defined to return the $\nonterm{tc}$ and $\nonterm{ec}$
-from the right-hand sides above respectively.
+from the respective right-hand sides.
 

### 2020-07-25 tex/paper.tex
Define \Tranterm for T-Abs

@@
     Thus, we are allowed to use the (derived) core typing rule \textsc{CT-VarW}.
+
+  \item \textsc{T-Abs}.
+    \begin{align*}
+      & \inferrule{\nonterm G \vdash (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2 \\\\
+                   \nonterm G , \gterm x : \nonterm{ts}_1 \vdash \nonterm{es} : \nonterm{ts}_2}
+                  {\nonterm G \vdash \lambda x : \nonterm{ts}_1. \nonterm{es} : (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2}
+          \longmapsto
+        \inferrule{\Tranty(\nonterm G \vdash (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2) \\\\
+                   \Tranterm(\nonterm G , \gterm x : \nonterm{ts}_1 \vdash \nonterm{es} : \nonterm{ts}_2)
+                   }
+                  {\nonterm G \vdash \lambda x : \nonterm{tc}_1. \nonterm{es} : \Pi \gterm x : \nonterm{tc}_1. \nonterm{tc}_2},    \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok),       \\
+                   &\nonterm{tc}_1 = \tranty(\Gamma \vdash \nonterm{tc}_1), \\
+                   &\nonterm{tc}_2 = \tranty(\Gamma , x : \nonterm{tc}_1 \vdash \nonterm{tc}_2), \\
+                   &\nonterm{ec} = \tranterm().
+    \end{align*}
+
+    The right-hand side here is well-formed.
+    Firstly, $\Tranty(\nonterm G \vdash (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2)$
+    produces a derivation tree of the form
+    $\nonterm{GC} \vdash \Pi \gterm x : \nonterm{tc}_1. \nonterm{tc}_2 : \star$,
+    since the only rule allowing deriving
+    $\nonterm G \vdash (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2$
+    is \textsc{TWF-Arr},
+    so it is sufficient to inspect the definition of $\Tranty$ for this rule.
+    Then, $\Tranterm(\nonterm G , \gterm x : \nonterm{ts}_1 \vdash \nonterm{es} : \nonterm{ts}_2)$
+    produces a derivation of $\nonterm{GC} , \gterm x : \nonterm{tc}_1 \vdash \nonterm{ec} : \nonterm{tc}_2$.
+    These two derivations can then be used as premises for \textsc{CT-Abs},
+    which is invoked on the right-hand side.
 \end{itemize}

### 2020-07-25 tex/paper.tex
Use bold font for \unit and \unitc
It's a constant after all, not a variable.
Update T-Unit description
\Tranctx doesn't map into well-formedness proofs of the context in the
core language.

@@
 
-\newcommand{\unit}{u}
+\newcommand{\unit}{\textbf{u}}
 \newcommand{\Unit}{\textbf{Unit}}
-\newcommand{\unitc}{\hat{u}}
+\newcommand{\unitc}{\hat{\textbf{u}}}
 \newcommand{\Unitc}{\widehat{\textbf{Unit}}}
@@
     This rule has a single premise $\nonterm G \ctxok$ that the surface context is well-formed.
-    We use $\Tranctx$ to translate the surface derivation of that premise into the core one,
-    and extract the resulting context $\nonterm{GC}$.
-    We then use that context for \textsc{CT-UnitTerm}.
+    $\Tranctx$ is used to translate the surface derivation of that premise into the core context $\nonterm{GC}$,
+    which is then used in \textsc{CT-UnitTerm}.
 

### 2020-07-25 tex/paper.tex
More styling fixes
Less "we", more passive voice.

@@
 and the rules for our extensions are defined in \figref{core_typing_exts}.
-Note that the \textsc{CT-ADTCase} rule uses the equality $\equiv$,
+Note that the \textsc{CT-ADTCase} rule uses $\equiv$ denoting the equality
 which is a derived form defined later.
@@
 Since $\Tranctx$ is defined on the derivations of $\nonterm G \ctxok$,
-we consider the last rule used in a derivation:
+consider the last rule used in a derivation:
 \begin{itemize}
@@
     \end{align*}
-    The base case is trivial: we map the empty context to the empty context.
+    The base case is trivial: the empty context is mapped to the empty context.
   \item \textsc{TCTX-Bind}.
@@
 
-More formally, we consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$:
+More formally, consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$:
 \begin{itemize}
@@
 
-    We do not mention the premises of the translated rule.
     By \cref{lma:non_dep_pair_typing}
@@
     Thus, there exists a derivation of $\nonterm{GC} \vdash (\hat B, \unitc \equiv \unitc) : \star$,
-    and it is effectively constructible~--- we just omit it for brevity.
+    and it is effectively constructible, although omitted for brevity.
   \item \textsc{TWF-Base}.
@@
 
-    We resort to \textsc{CT-Form} on the right.
-    This is well-formed,
+    The rule \textsc{CT-Form} is used on the right.
+    This is admissible,
     since $\Tranty(\nonterm G \vdash \nonterm{ts}_1)$ produces a derivation of $\nonterm{GC} \vdash \nonterm{tc}_1 : \star$,
-    and similarly we get a derivation of $\nonterm{GC}, \gterm x : \nonterm{tc}_1 \vdash \nonterm{tc}_2 : \star$ from
-    the other invocation of $\Tranty$.
+    and similarly there is a derivation of $\nonterm{GC}, \gterm x : \nonterm{tc}_1 \vdash \nonterm{tc}_2 : \star$
+    resulting from the other invocation of $\Tranty$.
   \item \textsc{TWF-ADT}.
@@
     To show that the right-hand side is well-defined,
-    we need to show that $x : \nonterm{tc} \in \nonterm{GC}$ holds.
+    it suffices to show that $x : \nonterm{tc} \in \nonterm{GC}$ holds.
 

### 2020-07-25 tex/paper.tex
Fix the stuff in T-Abs translation
I was tired and probalby should've gone to sleep.

@@
                    }
-                  {\nonterm G \vdash \lambda x : \nonterm{tc}_1. \nonterm{es} : \Pi \gterm x : \nonterm{tc}_1. \nonterm{tc}_2},    \\
+                  {\nonterm{GC} \vdash \lambda x : \nonterm{tc}_1. \nonterm{ec} : \Pi \gterm x : \nonterm{tc}_1. \nonterm{tc}_2},    \\
       \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok),       \\
-                   &\nonterm{tc}_1 = \tranty(\Gamma \vdash \nonterm{tc}_1), \\
-                   &\nonterm{tc}_2 = \tranty(\Gamma , x : \nonterm{tc}_1 \vdash \nonterm{tc}_2), \\
-                   &\nonterm{ec} = \tranterm().
+                   &\nonterm{tc}_1 = \tranty(\Gamma \vdash \nonterm{ts}_1), \\
+                   &\nonterm{tc}_2 = \tranty(\Gamma , x : \nonterm{ts}_1 \vdash \nonterm{ts}_2), \\
+                   &\nonterm{ec} = \tranterm(\nonterm G , \gterm x : \nonterm{ts}_1 \vdash \nonterm{es} : \nonterm{ts}_2).
     \end{align*}

### 2020-07-25 tex/paper.tex
Write down T-App translation

@@
     which is invoked on the right-hand side.
+
+  \item \textsc{T-App}.
+    \begin{align*}
+      & \inferrule{\nonterm G \vdash \nonterm{es}_1 : (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2 \\\\
+                   \nonterm G \vdash \nonterm{es}_2 : \nonterm{ts}_1}
+                  {\nonterm G \vdash \nonterm{es}_1 \nonterm{es}_2 : [ x \mapsto \nonterm{es}_2 ] \nonterm{ts}_2}
+          \longmapsto
+        \inferrule{\Tranterm(\nonterm G \vdash \nonterm{es}_1 : (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2) \\\\
+                   \Tranterm(\nonterm G \vdash \nonterm{es}_2 : \nonterm{ts}_1)
+                   }
+                  {\nonterm{GC} \vdash \nonterm{ec}_1 \nonterm{ec}_2 : [ x \mapsto \nonterm{ec}_2 ] \nonterm{tc}_2},    \\
+      \text{where}~&\nonterm{GC} = \Tranctx(\nonterm G \ctxok),       \\
+                   &\nonterm{tc}_2 = \tranty(\Gamma \vdash \nonterm{ts}_2), \\
+                   &\nonterm{ec}_1 = \tranterm(\nonterm G \vdash \nonterm{es}_1 : (\gterm x : \nonterm{ts}_1) \rightarrow \nonterm{ts}_2), \\
+                   &\nonterm{ec}_2 = \tranterm(\nonterm G \vdash \nonterm{es}_2 : \nonterm{ts}_1).
+    \end{align*}
+
+    Admissibility of the right-hand side follows from a similar argument.
 \end{itemize}

### 2020-07-25 tex/build.sh
Split ott defs into common, surface and core

@@ -2,5 +2,5 @@
 
 set -e
 
-ott -tex_wrap false -i surface.ott -o surface.ott.tex
+ott -tex_wrap false -i common.ott -i surface.ott -i core.ott -o surface.ott.tex
 pdflatex paper.tex && bibtex paper.aux && pdflatex paper.tex && pdflatex paper.tex

### 2020-07-25 tex/paper.tex
Add values to the syntax
Add \evalsto command and terminal
Surface language operational semantics
Also write a few words about the surface language operational semantics

@@
 \newcommand{\fresh}{\text{~fresh name}}
+\newcommand{\evalsto}{\rightsquigarrow}
 
@@
 \newnonterm{GC}{\hat\Gamma}
+\newnonterm{vs} \varpi
 
@@
 The syntax of the surface language is presented in \figref{surface_syntax}.
-
 The typing rules for the surface language are laid out in \figref{surface_typing}.
@@
 stating that the context is well-formed.
+The (small-step) operational semantics for the surface language are presented in \figref{surface_opsem}.
+That is largely the usual call-by-value evaluation model
+with some extra rules for the ADTs and pattern matching.
 
@@
   \begin{subfigure}{.6\textwidth}
-    \nonterms{es,l}
+    \nonterms{es,l,vs}
     \caption{Term level}
@@
   \label{fig:surface_typing}
+\end{figure}
+
+\begin{figure}[ht]
+  \footnotesize
+  \drules[E]{$\nonterm{es}_1 \evalsto \nonterm{es}_2$}{term evaluation}{AppL,AppR,AppAbs,ADT,CaseScrut,CaseMatch}
+  \caption{Surface language small-step operational semantics}
+  \label{fig:surface_opsem}
 \end{figure}

### 2020-07-26 tex/paper.tex
Core language values
Add most of core language evaluation rules
Mention the core evaluation rules in the text

@@
 \newnonterm{vs} \varpi
+\newnonterm{vc}{\hat\varpi}
 
@@
 
+Even though we believe that our core language still enjoys the strong normalization property,
+we still explicitly define the evaluation rules (\figref{core_opsem}).
+Firstly, this allows fixing a specific evaluation strategy mirroring the one for the surface language,
+simplifying the subsequent proofs of some key theorems.
+Secondly, the rule for \textsc{CE-CaseMatch},
+as well as our flavour of pattern matching in general,
+deserves some special attention.
+\todo[inline]{Pay said attention.}
+
 \begin{figure}[ht]
   \footnotesize
-  \nonterms{ec,s}
+  \nonterms{ec,s,vc}
   \caption{Core language syntax}
@@
   \label{fig:core_typing_exts}
+\end{figure}
+
+\begin{figure}[ht]
+  \footnotesize
+  \drules[CE]{$\nonterm{ec}_1 \evalsto \nonterm{ec}_2$}{term evaluation}{AppL,AppR,AppAbs,ADT,CaseScrut,CaseMatch}
+  \caption{Core language small-step operational semantics}
+  \label{fig:core_opsem}
 \end{figure}

### 2020-07-27 tex/build.sh
Change the build to allow for tex filtering via ott

@@ -2,5 +2,5 @@
 
 set -e
 
-ott -tex_wrap false -i common.ott -i surface.ott -i core.ott -o surface.ott.tex
+ott -tex_wrap false -i common.ott -i surface.ott -i core.ott -o surface.ott.tex -tex_filter paper.mng paper.tex
 pdflatex paper.tex && bibtex paper.aux && pdflatex paper.tex && pdflatex paper.tex

### 2020-07-27 tex/build.sh
Add some support for writing translations as ott expressions

@@ -2,5 +2,5 @@
 
 set -e
 
-ott -tex_wrap false -i common.ott -i surface.ott -i core.ott -o surface.ott.tex -tex_filter paper.mng paper.tex
+ott -tex_wrap false -i common.ott -i surface.ott -i core.ott -i translation.ott -o surface.ott.tex -tex_filter paper.mng paper.tex
 pdflatex paper.tex && bibtex paper.aux && pdflatex paper.tex && pdflatex paper.tex
